### 2021-06-07 paper.Rmd
infra structure in place - dummy example file rendered via rticles::ieee_article

---
title: "Untitled"
author: "PRU"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_book:
     base_format: rticles::ieee_article
abstract: |
  Our great abstract
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2021-06-07 paper.tex
infra structure in place - dummy example file rendered via rticles::ieee_article

%% template.tex
%% from
%% bare_conf.tex
%% V1.4b
%% 2015/08/26
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
%% conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/pkg/ieeetran
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE!
%% User assumes all risk.
%% In no event shall the IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. The IEEE's font choices and paper sizes can   ***
% *** trigger bugs that do not appear when using other class files.       ***                          ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

\documentclass[conference,final,]{IEEEtran}
% Some Computer Society conferences also require the compsoc mode option,
% but others use the standard conference format.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/pkg/ifpdf
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of the IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off
% such as if a citation ever needs to be enclosed in parenthesis.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 5.0 (2009-03-20) and later if using hyperref.sty.
% The latest version can be obtained at:
% http://www.ctan.org/pkg/cite
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at:
% http://www.ctan.org/pkg/graphicx
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/pkg/epslatex
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics.
%
% Note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/amsmath

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/pkg/algorithms
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/pkg/algorithmicx

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/array

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

% *** SUBFIGURE PACKAGES ***
%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false,font=footnotesize]{subfig}
%\fi
% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
% for subfigure.sty, the latter of which is no longer maintained and is
% incompatible with some LaTeX packages including fixltx2e. However,
% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
% which will override IEEEtran.cls' handling of captions and this will result
% in non-IEEE style figure/table captions. To prevent this problem, be sure
% and invoke subfig.sty's "caption=false" package option (available since
% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
% handling of captions.
% Note that the Computer Society format requires a larger sans serif font
% than the serif footnote size font used in traditional IEEE formatting
% and thus the need to invoke different subfig.sty package options depending
% on whether compsoc mode has been enabled.
%
% The latest version and documentation of subfig.sty can be obtained at:
% http://www.ctan.org/pkg/subfig

% *** FLOAT PACKAGES ***
%

%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure.
% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
% corrections already built into the system in which case a warning will
% be issued if an attempt is made to load fixltx2e.sty as it is no longer
% needed.
% The latest version and documentation can be found at:
% http://www.ctan.org/pkg/fixltx2e

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/stfloats
% Do not use the stfloats baselinefloat ability as the IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that the IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
% \usepackage{dblfloatfix}
% The latest version can be found at:
% http://www.ctan.org/pkg/dblfloatfix

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/url
% Basically, \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

%% BEGIN MY ADDITIONS %%

\usepackage{longtable,booktabs}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}

\usepackage[unicode=true]{hyperref}

\hypersetup{
            pdftitle={Untitled},
            pdfauthor={PRU},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls

% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{5}

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% Pandoc citation processing

% Pandoc header

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%% END MY ADDITIONS %%

\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Untitled}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations

\author{
}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
%
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3},
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Our great abstract
\end{abstract}

% keywords

% use for special paper notices

% make the title area
\maketitle

% no keywords

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\hypertarget{r-markdown}{%
\subsection{R Markdown}\label{r-markdown}}

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see \url{http://rmarkdown.rstudio.com}.

When you click the \textbf{Knit} button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{summary}\NormalTok{(cars)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      speed           dist
##  Min.   : 4.0   Min.   :  2.00
##  1st Qu.:12.0   1st Qu.: 26.00
##  Median :15.0   Median : 36.00
##  Mean   :15.4   Mean   : 42.98
##  3rd Qu.:19.0   3rd Qu.: 56.00
##  Max.   :25.0   Max.   :120.00
\end{verbatim}

\hypertarget{including-plots}{%
\subsection{Including Plots}\label{including-plots}}

You can also embed plots, for example:

\includegraphics{paper_files/figure-latex/pressure-1.pdf}

Note that the \texttt{echo\ =\ FALSE} parameter was added to the code chunk to prevent printing of the R code that generated the plot.

\end{document}

### 2021-06-07 R/set_proxy.R

@@ -1,0 +1,13 @@
+#' Utility function to help when password changed or the Environmental variables are not loaded
+#'
+#' @param .pw (string) password
+#' @param .name (string) username
+#'
+#' @return NIL sets proxy variables
+#' @export
+
+set_proxy <- function(.pw, .name="rkoelle"){
+  proxy_string <- paste0(.name,":",.pw,"@pac.eurocontrol.int:9512/M")
+  Sys.setenv("https_proxy" = paste0("https://", proxy_string))
+  Sys.setenv("http_proxy"  = paste0("http://",  proxy_string))
+}

### 2021-06-08 paper.Rmd
paper skeleton in place, infra with bib, etc. hooked up - ready to go

@@
 ---
-title: "Untitled"
-author: "PRU"
+title: "Assessing the Global COVID19 Impact on Air Transport with Open Data"
 date: "`r format(Sys.Date(), '%d %B %Y')`"
@@
      base_format: rticles::ieee_article
+bibliography: "DASC2021.bib"
+affiliation:
+  ## use one only of the following
+  # author-columnar: true         ## one column per author
+  institution-columnar: true  ## one column per institution  
+  # wide: true          ## one column wide author/affiliation fields
+  institution:
+    - name: EUROCONTROL
+      department: Performance Review Unit
+      location: Brussels (Belgium)
+      #mark: 1
+      author:
+        - name: Rainer Koelle
+          email: rainer.koelle@eurocontrol.int
+    - name: DECEA
+      department: tbd ... 
+      location: tbd location - suggest Rio (Brazil)
+      #mark: 2
+      author:
+        - name: Fabio Coneiro Barbarosa
+          email: second_author@somedomain.com
 abstract: |
-  Our great abstract
+  This paper approaches the impact of the pandemic as a massive service disruption of the pre-pandemic global     connectivity and regional air transport networks.
+In particular, the project aims to provide data analytical evidence for policy success and transformation of the air transportation system.
+As an aspirational goal, the industry aims to recover in a “greener” manner.
+The project builds on openly available data sets.
+The paper will be produced in a reproducible manner making the data, code, and its processing available to interested reseachers and practitioners.
+The open assessment will provide policy makers with a tool to assess the reaction to local or regional measures.
 ---
@@
 
-## R Markdown
+# Introduction
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+<!--| 1. Nature of the problem. -->
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+This paper is heavily informed by the work of [@strohmeier_crowdsourced_2021].
 
-```{r cars}
-summary(cars)
+For many years, many concerns of the global air traffic management community has been directed to the evident problem of imbalances between capacity and demand.
+The pressing, increasing demand for air transport registered in the last decade not only has already produced challenging delay management practices, but also fostered projections of even worse scenarios.
+EUROCONTROL (\_\_\_\_), for example, argued that delays in Europe could reach up to 20 minutes per flight in 2040, in stark contrast to the 12 minutes per flight, as registered in 2016.
+
+In the above scenario, many disturbances on the air navigation system could represent a real threat to multiple stakeholders.
+Events such as extreme bad weather, unexpected interruptions of air navigation services, changes in regulatory framework and others: all of those inputs could promote even more delay and its propagation effects.
+That is why the concept of resilience in ATM system became similarly relevant in the agenda during the same period.
+Arguably, a resilient ATM system could mitigate the negative effects of excessive demands on insufficient capacity and their respective constraints and bottlenecks.
+
+However, the recent COVID-19 crisis posed a completely different, unexpected, and inverted challenge.
+Demand for air transport dropped as low as 90% of the previous "normal" in many places.
+Where the lack of capacity was previously the issue, now the lack of demand threatened the ATM system stability.
+In the financial perspective, airlines and airports had to deal with an unprecedented decrease in incomes.
+As a result, air navigation providers collected less fees for their services, due to significantly fewer flights.
+In the operational perspective, pilots and air traffic controllers practiced less.
+The problems and obstacles developed into many other dimensions.
+
+Hence, the current scenario is a proper moment to further investigate the concept of resilience.
+
+```{r}
+# Problem Statement
+
+# The problem is that, currently, the concept
+# of resilience is mostly directed to recovery
+# against delay propagation after negative 
+# disturbances. However, the current scenario
+# poses an inverted challenge, of very low 
+# delays due to low demand against surplus 
+# capacity. Therefore, there is room for 
+# enlarging the comprehension of the concept
+# of resilience in ATM systems.
+
+# # Purpose Statement
+# 
+# ???The purpose of this research is to 
+# investigate additional dimensions in which 
+# resilience could be measured, in addition
+# to the current framework of delay analysis.
+# 
+# Research Question
+# 
+# ???How can we enlarge the concept of 
+# resilience, so that it is applicable to
+# scenarios of low traffic? 
 ```
 
-## Including Plots
+<!--| 2. High-Level background of previous work or relevance. -->
 
-You can also embed plots, for example:
+<!--| 3. Purpose and significance of the paper.  -->
 
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
+<!--| 4. Method by which the problem is approached. -->
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+<!-- 5. Contribution of paper -->
+
+The contribution of this paper are
+
+-   conceptualisation of the COVID-19 impact on air transportation as a resilience problem;
+
+-   assessing the impact on the basis of open data <!-- specify a bit more what we do/model/analyse -->
+
+-   identification of patterns and/or measures to describe and quantify/evaluate the level of recovery (or disruption)
+
+# Background
+
+## COVID-19 & Air Transportation
+
+## Resilience
+
+EUROCONTROL (2009): first definition of resilience in ATM context -- "Resilience is the intrinsic ability of a system to adjust its functioning prior to, during, or following changes and disturbances, so that it can sustain required operations under both expected and unexpected conditions".
+
+Gluchshenko (2012):
+
+Definitions for Resilience, robustness, disturbance, stress, and perturbation Proposition for a framework of different levels of stress/perturbations Proposition of metrics for resilience (both quantitative and qualitative)
+
+Gluchshenko (2013): repeats the previous ideas and adds a performance-based approach as well as an algorithm to investigate resilience
+
+Project Resilience 2050 (Jun/2012 + 43 months) -- includes the previous definitions and other technical tasks.
+However, it evolves the way to measure resilience.
+Now, not only the time of deviation and time of recovery is considered.
+The project measures it as the relative difference of rate of delays correlation, or R = (ax1 -- dx1)/dx1 -- it has no unit, it's the difference between two pearson correlations.
+
+Koelle (2015): proposes to address resilience as a situation management and state-oriented problem.
+Through two case studies, argued that "there is a lack of fit of the current operational ANS performance indicators to address impact of disruptions as they are primarily based on actual timestamps or transition times."
+
+## \<if we need to fill space\> Crowd-Sourced Data Collection
+
+<!-- provide overview of data collection through crowd-sourced network and communities -->
+
+# Method/Materials
+
+## Open-source Data
+
+# Results/Discussion
+
+to be developed
+
+# Conclusion
+
+<!-- 1. What is shown by this work and its significance.  -->
+
+<!-- 2. Limitations and advantages. -->
+
+<!-- 3. Applications of the results. -->
+
+<!-- 4. Recommendations for further work. -->
+
+# Acknowledgment {.unnumbered}
+
+# References {.unnumbered}
+
+<!-- DO NOT EDIT BELOW THIS POINT ----------------------------------->

### 2021-06-08 paper.tex
paper skeleton in place, infra with bib, etc. hooked up - ready to go

@@
 \usepackage{longtable,booktabs}
-\usepackage{graphicx}
-% We will generate all images so they have a width \maxwidth. This means
-% that they will get their normal width if they fit onto the page, but
-% are scaled down if they would overflow the margins.
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
-\else\Gin@nat@width\fi}
-\makeatother
-\let\Oldincludegraphics\includegraphics
-\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
 
@@
 \hypersetup{
-            pdftitle={Untitled},
-            pdfauthor={PRU},
+            pdftitle={Assessing the Global COVID19 Impact on Air Transport with Open Data},
             pdfborder={0 0 0},
@@
 % Pandoc citation processing
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc} % for calculating minipage widths
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 % Do not put math or special symbols in the title.
-\title{Untitled}
+\title{Assessing the Global COVID19 Impact on Air Transport with Open Data}
 
@@
 \author{
+
+%% ---- classic IEEETrans wide authors' list ----------------
+ % -- end affiliation.wide
+%% ----------------------------------------------------------
+
+%% ---- classic IEEETrans one column per institution --------
+ %% -- beg if/affiliation.institution-columnar
+\IEEEauthorblockN{
+  %% -- beg for/affiliation.institution.author
+Rainer Koelle %% -- end for/affiliation.institution.author
+}
+\IEEEauthorblockA{Performance Review Unit\\
+EUROCONTROL\\
+Brussels (Belgium)
+  %% -- beg for/affiliation.institution.author
+\\rainer.koelle@eurocontrol.int
+ %% -- end for/affiliation.institution.author
+}
+\and
+\IEEEauthorblockN{
+  %% -- beg for/affiliation.institution.author
+Fabio Coneiro Barbarosa %% -- end for/affiliation.institution.author
+}
+\IEEEauthorblockA{tbd \ldots{}\\
+DECEA\\
+tbd location - suggest Rio (Brazil)
+  %% -- beg for/affiliation.institution.author
+\\second\_author@somedomain.com
+ %% -- end for/affiliation.institution.author
+}
+ %% -- end for/affiliation.institution
+ %% -- end if/affiliation.institution-columnar
+%% ----------------------------------------------------------
+
+%% ---- one column per author, classic/default IEEETrans ----
+ %% -- end if/affiliation.institution-columnar
+%% ----------------------------------------------------------
+
 }
@@
 \begin{abstract}
-Our great abstract
+This paper approaches the impact of the pandemic as a massive service disruption of the pre-pandemic global connectivity and regional air transport networks.
+In particular, the project aims to provide data analytical evidence for policy success and transformation of the air transportation system.
+As an aspirational goal, the industry aims to recover in a ``greener'' manner.
+The project builds on openly available data sets.
+The paper will be produced in a reproducible manner making the data, code, and its processing available to interested reseachers and practitioners.
+The open assessment will provide policy makers with a tool to assess the reaction to local or regional measures.
 \end{abstract}
@@
 
-\hypertarget{r-markdown}{%
-\subsection{R Markdown}\label{r-markdown}}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see \url{http://rmarkdown.rstudio.com}.
+This paper is heavily informed by the work of (Strohmeier et al. 2021).
 
-When you click the \textbf{Knit} button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+For many years, many concerns of the global air traffic management community has been directed to the evident problem of imbalances between capacity and demand.
+The pressing, increasing demand for air transport registered in the last decade not only has already produced challenging delay management practices, but also fostered projections of even worse scenarios.
+EUROCONTROL (\_\_\_\_), for example, argued that delays in Europe could reach up to 20 minutes per flight in 2040, in stark contrast to the 12 minutes per flight, as registered in 2016.
+
+In the above scenario, many disturbances on the air navigation system could represent a real threat to multiple stakeholders.
+Events such as extreme bad weather, unexpected interruptions of air navigation services, changes in regulatory framework and others: all of those inputs could promote even more delay and its propagation effects.
+That is why the concept of resilience in ATM system became similarly relevant in the agenda during the same period.
+Arguably, a resilient ATM system could mitigate the negative effects of excessive demands on insufficient capacity and their respective constraints and bottlenecks.
+
+However, the recent COVID-19 crisis posed a completely different, unexpected, and inverted challenge.
+Demand for air transport dropped as low as 90\% of the previous ``normal'' in many places.
+Where the lack of capacity was previously the issue, now the lack of demand threatened the ATM system stability.
+In the financial perspective, airlines and airports had to deal with an unprecedented decrease in incomes.
+As a result, air navigation providers collected less fees for their services, due to significantly fewer flights.
+In the operational perspective, pilots and air traffic controllers practiced less.
+The problems and obstacles developed into many other dimensions.
+
+Hence, the current scenario is a proper moment to further investigate the concept of resilience.
 
@@
 \begin{Highlighting}[]
-\FunctionTok{summary}\NormalTok{(cars)}
+\CommentTok{\# Problem Statement }
+
+\CommentTok{\# The problem is that, currently, the concept}
+\CommentTok{\# of resilience is mostly directed to recovery}
+\CommentTok{\# against delay propagation after negative }
+\CommentTok{\# disturbances. However, the current scenario}
+\CommentTok{\# poses an inverted challenge, of very low }
+\CommentTok{\# delays due to low demand against surplus }
+\CommentTok{\# capacity. Therefore, there is room for }
+\CommentTok{\# enlarging the comprehension of the concept}
+\CommentTok{\# of resilience in ATM systems. }
+
+\CommentTok{\# \# Purpose Statement}
+\CommentTok{\# }
+\CommentTok{\# ???The purpose of this research is to }
+\CommentTok{\# investigate additional dimensions in which }
+\CommentTok{\# resilience could be measured, in addition}
+\CommentTok{\# to the current framework of delay analysis.}
+\CommentTok{\# }
+\CommentTok{\# Research Question}
+\CommentTok{\# }
+\CommentTok{\# ???How can we enlarge the concept of }
+\CommentTok{\# resilience, so that it is applicable to}
+\CommentTok{\# scenarios of low traffic? }
 \end{Highlighting}
@@
 
-\begin{verbatim}
-##      speed           dist
-##  Min.   : 4.0   Min.   :  2.00
-##  1st Qu.:12.0   1st Qu.: 26.00
-##  Median :15.0   Median : 36.00
-##  Mean   :15.4   Mean   : 42.98
-##  3rd Qu.:19.0   3rd Qu.: 56.00
-##  Max.   :25.0   Max.   :120.00
-\end{verbatim}
+The contribution of this paper are
 
-\hypertarget{including-plots}{%
-\subsection{Including Plots}\label{including-plots}}
+\begin{itemize}
+\item
+  conceptualisation of the COVID-19 impact on air transportation as a resilience problem;
+\item
+  assessing the impact on the basis of open data
+\item
+  identification of patterns and/or measures to describe and quantify/evaluate the level of recovery (or disruption)
+\end{itemize}
 
-You can also embed plots, for example:
+\hypertarget{background}{%
+\section{Background}\label{background}}
 
-\includegraphics{paper_files/figure-latex/pressure-1.pdf}
+\hypertarget{covid-19-air-transportation}{%
+\subsection{COVID-19 \& Air Transportation}\label{covid-19-air-transportation}}
 
-Note that the \texttt{echo\ =\ FALSE} parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+\hypertarget{resilience}{%
+\subsection{Resilience}\label{resilience}}
+
+EUROCONTROL (2009): first definition of resilience in ATM context -- ``Resilience is the intrinsic ability of a system to adjust its functioning prior to, during, or following changes and disturbances, so that it can sustain required operations under both expected and unexpected conditions.''
+
+Gluchshenko (2012):
+
+Definitions for Resilience, robustness, disturbance, stress, and perturbation Proposition for a framework of different levels of stress/perturbations Proposition of metrics for resilience (both quantitative and qualitative)
+
+Gluchshenko (2013): repeats the previous ideas and adds a performance-based approach as well as an algorithm to investigate resilience
+
+Project Resilience 2050 (Jun/2012 + 43 months) -- includes the previous definitions and other technical tasks.
+However, it evolves the way to measure resilience.
+Now, not only the time of deviation and time of recovery is considered.
+The project measures it as the relative difference of rate of delays correlation, or R = (ax1 -- dx1)/dx1 -- it has no unit, it's the difference between two pearson correlations.
+
+Koelle (2015): proposes to address resilience as a situation management and state-oriented problem.
+Through two case studies, argued that ``there is a lack of fit of the current operational ANS performance indicators to address impact of disruptions as they are primarily based on actual timestamps or transition times.''
+
+\hypertarget{if-we-need-to-fill-space-crowd-sourced-data-collection}{%
+\subsection{\textless if we need to fill space\textgreater{} Crowd-Sourced Data Collection}\label{if-we-need-to-fill-space-crowd-sourced-data-collection}}
+
+\hypertarget{methodmaterials}{%
+\section{Method/Materials}\label{methodmaterials}}
+
+\hypertarget{open-source-data}{%
+\subsection{Open-source Data}\label{open-source-data}}
+
+\hypertarget{resultsdiscussion}{%
+\section{Results/Discussion}\label{resultsdiscussion}}
+
+to be developed
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+\hypertarget{acknowledgment}{%
+\section*{Acknowledgment}\label{acknowledgment}}
+\addcontentsline{toc}{section}{Acknowledgment}
+
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-strohmeier_crowdsourced_2021}{}%
+Strohmeier, Martin, Xavier Olive, Jannis Lübbe, Matthias Schäfer, and Vincent Lenders.
+2021.
+{``Crowdsourced Air Traffic Data from {OpenSky Network} 2019-2020.''} \emph{Earth Systems Science Data} 13: 357--66.
+\url{https://doi.org/10.5194/essd-13-357-2021}.
+
+\end{CSLReferences}
 

### 2021-06-09 paper.Rmd
check points for abstract inserted (as javascript comments)
a few breadcrumbs

@@
 The open assessment will provide policy makers with a tool to assess the reaction to local or regional measures.
+  <!-- motivation -->
+  <!-- short background - why important -->
+  <!-- approach chosen / research method -->
+  <!-- experimental work -->
+  <!-- results obtained -->
+  <!-- conclusions / key take-away --> 
 ---
@@
 <!--| 3. Purpose and significance of the paper.  -->
+This paper approaches the impact of the pandemic as a massive service disruption of the pre-pandemic global connectivity and regional air transport networks.
+In particular, the project aims to provide data analytical evidence for policy success and transformation of the air transportation system.
+As an aspirational goal, the industry aims to recover in a “greener” manner.
+To date, no assessment of this transformational aspects has been conducted.
 
 <!--| 4. Method by which the problem is approached. -->
+
+* data-analytical approach - using open data / freely available (tbd: validated against organisational data)
+* ...
 

### 2021-06-09 paper.tex
a few breadcrumbs

@@
 
+This paper approaches the impact of the pandemic as a massive service disruption of the pre-pandemic global connectivity and regional air transport networks.
+In particular, the project aims to provide data analytical evidence for policy success and transformation of the air transportation system.
+As an aspirational goal, the industry aims to recover in a ``greener'' manner.
+To date, no assessment of this transformational aspects has been conducted.
+
+\begin{itemize}
+\tightlist
+\item
+  data-analytical approach - using open data / freely available (tbd: validated against organisational data)
+\item
+  \ldots{}
+\end{itemize}
+
 The contribution of this paper are

### 2021-06-09 paper.Rmd
Updated header info

@@
     - name: DECEA
-      department: tbd ... 
-      location: tbd location - suggest Rio (Brazil)
+      department: Subdepartment of Operations 
+      location: Rio de Janeiro (Brazil)
       #mark: 2
       author:
-        - name: Fabio Coneiro Barbarosa
-          email: second_author@somedomain.com
+        - name: Fabio Lourenco Carneiro Barbosa
+          email: barbosaflcb@fab.mil.br
 abstract: |

### 2021-06-09 paper.Rmd
It worked!! TinyTex fixed!

@@
 <!--| 3. Purpose and significance of the paper.  -->
+
 This paper approaches the impact of the pandemic as a massive service disruption of the pre-pandemic global connectivity and regional air transport networks.
 In particular, the project aims to provide data analytical evidence for policy success and transformation of the air transportation system.
-As an aspirational goal, the industry aims to recover in a “greener” manner.
+As an aspirational goal, the industry aims to recover in a "greener" manner.
 To date, no assessment of this transformational aspects has been conducted.
@@
 
-* data-analytical approach - using open data / freely available (tbd: validated against organisational data)
-* ...
+-   data-analytical approach - using open data / freely available (tbd: validated against organisational data)
+-   ...
 

### 2021-06-09 paper.tex
It worked!! TinyTex fixed!

@@
   %% -- beg for/affiliation.institution.author
-Fabio Coneiro Barbarosa %% -- end for/affiliation.institution.author
+Fabio Lourenco Carneiro Barbosa %% -- end for/affiliation.institution.author
 }
-\IEEEauthorblockA{tbd \ldots{}\\
+\IEEEauthorblockA{Subdepartment of Operations\\
 DECEA\\
-tbd location - suggest Rio (Brazil)
+Rio de Janeiro (Brazil)
   %% -- beg for/affiliation.institution.author
-\\second\_author@somedomain.com
+\\barbosaflcb@fab.mil.br
  %% -- end for/affiliation.institution.author

### 2021-06-15 paper.Rmd
initital stab

@@
 knitr::opts_chunk$set(echo = TRUE)
+library(tidyverse)
+library(lubridate)
+
+```
+
+```{r}
+#CURRENT COMMENTS AND TO-DO'S
+# I HAVE DOWNLOADED 3 FILES FOR NOW (APR/19, APR/2020, APR/2021), JUST TO START "TIDYING" AND EXPLORING.
+#IT'S IN THE DATA-RAW FOLDER (NOT SHARED WITH GITHUB), AS ALWAYS.
 ```
@@
 # scenarios of low traffic? 
+#
+# ???Research Question: 
+# ???RQ1.What was the impact of the pandemic on ATM resilience?
+# ???  RQ1.1 How resilience can be modeled in a low-demand scenario?
+# ???  RQ1.2 How resilient were different ATM systems worldwide?
+
 ```
@@
 -   data-analytical approach - using open data / freely available (tbd: validated against organisational data)
--   ...
+-   ???RQ1.1 = through a qualitative analysis of previous proposed models
+-   ???RQ1.2 = through a quantitative analysis of open data
 
@@
 
+A mixed-method approach, based on:
+
+  a) to answer RQ1.1, a qualitative analysis of previous models to develop acute low-demand as a disturbance
+  b) to answer RQ1.2, a quantitative analysis of open data, to observe (or not) different levels/stages of stress/recovery, which could indicate different "more" or "less" resilience to the disturbances
+
 ## Open-source Data
@@
 
-to be developed
+1.1
+
+a) Resilience can be measured as a function of time - the smaller the relationship between time of stress and the time of recovery, more resilient a system is.
+
+1.2 how to use open data to "see" resilience?
+
+1.2.1 Gather and prepare data
+```{r}
+#Reading and saving data
+source("./R/list_apt_files.R")
+file_names <- list_apt_files()
+open_sky <- map_dfr(file_names, read_csv)
+#write_csv(open_sky, "./data/open_sky.csv")
+
+# Take a look
+glimpse(open_sky)
+str(open_sky)
+summary(open_sky)
+
+# First exploration
+fb1 <- open_sky %>% transmute(typecode = as.factor(typecode), origin = as.factor(origin), destination = as.factor(destination), firstseen, lastseen)
+str(fb1)
+summary(fb1)
+# More than one million NA's
+
+fb2 <- fb1 %>% drop_na()
+str(fb2)
+summary(fb2)
+# Better...
+
+#Visualizing something...
+df1 <- fb2 %>% group_by(DATE = date(firstseen)) %>% summarize(count = n())
+df1 %>% ggplot(aes(x = as.factor(year(DATE)))) +
+  geom_boxplot(aes(y = count))
+
+```
+
+1.2.2
 

### 2021-06-15 R/list_apt_files.R

@@ -1,0 +1,12 @@
+#List files to be processed
+# .aprt is a 4-letter ICAO designator, set "*" for all
+# .year is a 4-digit year, set "*" for all
+# . opr is either "DEP" for departures or "ARR" for arrivals/landings
+
+list_apt_files <- function(.aprt = "*", .year = "*", .opr = "*", .path = "./data-raw/"){
+  ptn <- paste0(.aprt, ".*.", .year, ".*.", .opr)
+  fns <- list.files(.path, ptn)
+  full_fns <- c(paste0(.path, fns))
+  return(full_fns)
+}
+

### 2021-06-16 paper.Rmd
minors

@@
 #write_csv(open_sky, "./data/open_sky.csv")
+```
 
+```{r}
 # Take a look
@@
   geom_boxplot(aes(y = count))
+# This is the daily traffic of Aprils, distributed as a boxplot, FOR THE WHOLE WORLD! Needs to filter.
+
+#Filtering Brazilian and European data to reduce the sample (and csv) size
+#Preparing airports file
+
+bra_10_apts <- c("SBBR", "SBGR", "SBSP", "SBKP", "SBRJ", "SBGL", "SBCF", "SBSV", "SBPA", "SBCT")
+eur_apts <- c("EHAM","LFPG","EGLL","EDDF","EDDM","LEMD","LIRF","LEBL","EGKK","LSZH")
+study_airports <- c(bra_10_apts, eur_apts)
+study_airports
+
+fb2 <- fb2 %>% filter(origin %in% bra_10_apts | destination %in% bra_10_apts) %>% mutate(REG = "BRA", .before = typecode)
+fb2
+summary(fb2)
+
+#airports <- read_csv("data-raw/airports.csv")
+#airports <- airports %>% transmute(ICAO = ident, CONTINENT = continent, COUNTRY = iso_country, REGION = iso_region, NAME #= name, CITY = municipality) 
+#airports <- airports %>% mutate(CONTINENT = case_when(is.na(CONTINENT) ~ "NA", TRUE ~ CONTINENT))
+#airports
+#write_csv(airports, "./data/airports_info.csv")
 

### 2021-06-18 paper.Rmd
today's talk was excellent

@@
 ```{r}
-#Reading and saving data
+#Reading raw data protocol
+
 source("./R/list_apt_files.R")
-file_names <- list_apt_files()
+year <- "2021"
+file_names <- list_apt_files(.year = year)
 open_sky <- map_dfr(file_names, read_csv)
+fb1 <- open_sky %>% transmute(typecode = as.factor(typecode), origin = as.factor(origin), destination = as.factor(destination), firstseen, lastseen) %>% drop_na()
+
 #write_csv(open_sky, "./data/open_sky.csv")
+
+#Reading prepared data
+open_sky <- read_csv("data/open_sky.csv")
 ```
@@
 #Preparing airports file
+#Here we can control the selective sample
 
@@
 study_airports <- c(bra_10_apts, eur_apts)
-study_airports
 
-fb2 <- fb2 %>% filter(origin %in% bra_10_apts | destination %in% bra_10_apts) %>% mutate(REG = "BRA", .before = typecode)
-fb2
-summary(fb2)
+fb3 <- fb2 %>% filter(origin %in% study_airports | destination %in% study_airports)
+
+base_dataset <- fb3 %>% mutate(BRA_ORIGIN = origin %in% bra_10_apts,
+                        BRA_DEST = destination %in% bra_10_apts,
+                        EUR_ORIGIN = origin %in% eur_apts,
+                        EUR_DEST = destination %in% eur_apts,
+                        BRA_EUR_DIRECT = (origin %in% bra_10_apts | destination %in% bra_10_apts) & (origin %in% eur_apts | destination %in% eur_apts))
+
+# I opened a lot of logical variables to facilitate us knowing how many flights there are
+
+summary(base_dataset)
+
+temp1 <- base_dataset %>% group_by(
+  YR = year(firstseen),
+  ADEP = origin,
+  ADES = destination) %>%
+  summarise(COUNT = n(), .groups = "drop") %>%
+  arrange(desc(COUNT)) %>%
+  transmute(ROUTE = paste(ADEP, ADES, sep = "-"), YR, COUNT) %>%
+  group_by(YR) %>%
+  slice_head(n = 15)
+temp1
+
+temp1 %>% ggplot(aes(x = COUNT)) +
+  geom_col(aes(y = ROUTE)) +
+  facet_grid(.~YR)
 

### 2021-06-18 R/list_apt_files.R

@@ -3,8 +3,8 @@
 # .year is a 4-digit year, set "*" for all
 # . opr is either "DEP" for departures or "ARR" for arrivals/landings
 
-list_apt_files <- function(.aprt = "*", .year = "*", .opr = "*", .path = "./data-raw/"){
-  ptn <- paste0(.aprt, ".*.", .year, ".*.", .opr)
+list_apt_files <- function(.year = "*", .path = "./data-raw/"){
+  ptn <- paste0("flightlist_*.", .year, ".*.")
   fns <- list.files(.path, ptn)
   full_fns <- c(paste0(.path, fns))
   return(full_fns)

### 2021-06-18 paper.Rmd
up until worldwide data, without the regions yet

@@
 ```{r}
-#Reading raw data protocol
+#Reading raw data
 
 source("./R/list_apt_files.R")
-year <- "2021"
+year <- "202105"
 file_names <- list_apt_files(.year = year)
 open_sky <- map_dfr(file_names, read_csv)
-fb1 <- open_sky %>% transmute(typecode = as.factor(typecode), origin = as.factor(origin), destination = as.factor(destination), firstseen, lastseen) %>% drop_na()
 
-#write_csv(open_sky, "./data/open_sky.csv")
+#Selecting relevant variables
 
-#Reading prepared data
-open_sky <- read_csv("data/open_sky.csv")
+fb <- open_sky %>% transmute(ADEP = origin, ADES = destination, TYPE = as.factor(typecode), DATE = date(day), CALL = callsign, ACFT_ID = aircraft_uid)
+
+#Easily "dropping NA's# - this can be further sofisticated
+
+fb1 <- fb %>% drop_na()
+fb1
 ```
@@
 ```{r}
-# Take a look
-glimpse(open_sky)
-str(open_sky)
-summary(open_sky)
+#Joining coutries and regions
+# NOTE: the airports.csv file below is the downloaded file from www.ourairports.org.
 
-# First exploration
-fb1 <- open_sky %>% transmute(typecode = as.factor(typecode), origin = as.factor(origin), destination = as.factor(destination), firstseen, lastseen)
-str(fb1)
-summary(fb1)
-# More than one million NA's
+airports <- read_csv("data-raw/airports.csv") %>% transmute(ICAO = ident, CTRY = iso_country)
+#There are missing airports
+airports <- airports %>% add_row(ICAO  = c("SPJC", "YBMC", "LSZM", "YSCH", "EPLB", "K3M3", "VV01", "SC28", "CWF2", "EHDB", "74XA", "HE13"), CTRY  = c("PE", "AU", "FR", "AU", "PL", "US", "VN", "US", "CA", "NL", "US", "EG"))
 
-fb2 <- fb1 %>% drop_na()
-str(fb2)
-summary(fb2)
-# Better...
+#Joining to ADEP
+fb2 <- left_join(fb1, airports, by = c("ADEP" = "ICAO")) %>% mutate(ADEP_CTRY = CTRY, .keep = "unused", .after = ADEP)
+#Joining to ADES
+fb3 <- left_join(fb2, airports, by = c("ADES" = "ICAO")) %>% mutate(ADES_CTRY = CTRY, .keep = "unused", .after = ADES)
 
-#Visualizing something...
-df1 <- fb2 %>% group_by(DATE = date(firstseen)) %>% summarize(count = n())
-df1 %>% ggplot(aes(x = as.factor(year(DATE)))) +
-  geom_boxplot(aes(y = count))
-# This is the daily traffic of Aprils, distributed as a boxplot, FOR THE WHOLE WORLD! Needs to filter.
+#Check  NA's
+colSums(is.na(fb3))
+# Very few NA's - it's safe to drop and factor now
+
+world_dataset <- fb3 %>% drop_na() %>% mutate(ADEP_CTRY = as.factor(ADEP_CTRY), ADES_CTRY = as.factor(ADES_CTRY))
+world_dataset
+```
+
+```{r}
 

### 2021-06-18 paper.Rmd
good to loop

--- 2021-06-16 paper.Rmd
@@
 
+```
+
+```{r}
+#If someone needs filters below, here we can control any selective sample
+#Filtering Brazilian and European data to reduce the sample (and csv) size
+#Currently, those filters are not in use in the code, they are here just in case.
+
+bra_10_apts <- c("SBBR", "SBGR", "SBSP", "SBKP", "SBRJ", "SBGL", "SBCF", "SBSV", "SBPA", "SBCT")
+eur_apts <- c("EHAM","LFPG","EGLL","EDDF","EDDM","LEMD","LIRF","LEBL","EGKK","LSZH")
+study_airports <- c(bra_10_apts, eur_apts)
 ```
@@
 ```{r}
-#Reading and saving data
+#Reading raw data
+
 source("./R/list_apt_files.R")
-file_names <- list_apt_files()
+
+#Here I will assign only one month - "202105". If you want to include a full year, just assign year to "2021" or "2020". It works.
+year <- "202105"
+file_names <- list_apt_files(.year = year)
 open_sky <- map_dfr(file_names, read_csv)
-#write_csv(open_sky, "./data/open_sky.csv")
+
+#Selecting relevant variables
+
+fb <- open_sky %>% transmute(ADEP = origin, ADES = destination, TYPE = as.factor(typecode), DATE = date(day), CALL = callsign, ACFT_ID = aircraft_uid)
+
+#Easily "dropping NA's# - this can be further sofisticated
+
+fb1 <- fb %>% drop_na()
+fb1
 ```
@@
 ```{r}
-# Take a look
-glimpse(open_sky)
-str(open_sky)
-summary(open_sky)
+#Joining countries and regions
+# NOTE: the airports.csv file below is the downloaded file from www.ourairports.org.
 
-# First exploration
-fb1 <- open_sky %>% transmute(typecode = as.factor(typecode), origin = as.factor(origin), destination = as.factor(destination), firstseen, lastseen)
-str(fb1)
-summary(fb1)
-# More than one million NA's
+airports <- read_csv("data-raw/airports.csv") %>% mutate(continent = case_when(is.na(continent) ~ "NA", TRUE ~ continent))
+apt_countries <- airports %>% transmute(ICAO = ident, CTRY = iso_country)
 
-fb2 <- fb1 %>% drop_na()
-str(fb2)
-summary(fb2)
-# Better...
+#There are missing airports
+apt_countries <- apt_countries %>% add_row(ICAO  = c("SPJC", "YBMC", "LSZM", "YSCH", "EPLB", "K3M3", "VV01", "SC28", "CWF2", "EHDB", "74XA", "HE13"), CTRY  = c("PE", "AU", "FR", "AU", "PL", "US", "VN", "US", "CA", "NL", "US", "EG"))
 
-#Visualizing something...
-df1 <- fb2 %>% group_by(DATE = date(firstseen)) %>% summarize(count = n())
-df1 %>% ggplot(aes(x = as.factor(year(DATE)))) +
-  geom_boxplot(aes(y = count))
-# This is the daily traffic of Aprils, distributed as a boxplot, FOR THE WHOLE WORLD! Needs to filter.
+#If you need to write
+#write_csv(apt_countries, "./data/apt_countries.csv")
 
-#Filtering Brazilian and European data to reduce the sample (and csv) size
-#Preparing airports file
+#Joining to ADEP
+fb2 <- left_join(fb1, apt_countries, by = c("ADEP" = "ICAO")) %>% mutate(ADEP_CTRY = CTRY, .keep = "unused", .after = ADEP)
+#Joining to ADES
+fb3 <- left_join(fb2, apt_countries, by = c("ADES" = "ICAO")) %>% mutate(ADES_CTRY = CTRY, .keep = "unused", .after = ADES)
 
-bra_10_apts <- c("SBBR", "SBGR", "SBSP", "SBKP", "SBRJ", "SBGL", "SBCF", "SBSV", "SBPA", "SBCT")
-eur_apts <- c("EHAM","LFPG","EGLL","EDDF","EDDM","LEMD","LIRF","LEBL","EGKK","LSZH")
-study_airports <- c(bra_10_apts, eur_apts)
-study_airports
+#Check  NA's
+colSums(is.na(fb3))
+# Very few NA's - it's safe to drop and factor now
 
-fb2 <- fb2 %>% filter(origin %in% bra_10_apts | destination %in% bra_10_apts) %>% mutate(REG = "BRA", .before = typecode)
-fb2
-summary(fb2)
+fb4 <- fb3 %>% drop_na() %>% mutate(ADEP_CTRY = as.factor(ADEP_CTRY), ADES_CTRY = as.factor(ADES_CTRY))
 
-#airports <- read_csv("data-raw/airports.csv")
-#airports <- airports %>% transmute(ICAO = ident, CONTINENT = continent, COUNTRY = iso_country, REGION = iso_region, NAME #= name, CITY = municipality) 
-#airports <- airports %>% mutate(CONTINENT = case_when(is.na(CONTINENT) ~ "NA", TRUE ~ CONTINENT))
-#airports
-#write_csv(airports, "./data/airports_info.csv")
+# Associate the regions
+a <- airports %>% filter(continent == "EU") %>% select(iso_country) %>% unique()
+eur_countries <- a$iso_country
 
+#NOTE: Here we can adjust the European countries that comprises EU region by editing the "eur_countries" vector:
+# Currently in european countries: "GB" "AD" "ES" "AL" "XK" "AT" "BA" "BE" "BG" "IS" "BY" "UA" "CH" "CZ" "SK" "DE" "RU" "DK" "HR" "EE" "FI" "GG" "JE" "IM" "NL" "IE" "FO" "LU" "NO" "PL" "PT" "SE" "LV" "LT" "FR" "GR" "HU" "IT" "LI" "SI" "MT" "MC" "RO" "TR" "MD" "MK" "GI" "RS" "ME" "SM" "GE" "VA"
+
+base_dataset <- fb4 %>% mutate(ADEP_REG = as.factor(case_when(ADEP_CTRY == "US" ~ "US",
+                                    ADEP_CTRY == "BR" ~ "BR",
+                                    ADEP_CTRY %in% eur_countries ~ "EU",
+                                    TRUE ~ "Other")), .after = ADEP_CTRY) %>%
+                mutate(ADES_REG = as.factor(case_when(ADES_CTRY == "US" ~ "US",
+                                    ADES_CTRY == "BR" ~ "BR",
+                                    ADES_CTRY %in% eur_countries ~ "EU",
+                                    TRUE ~ "Other")), .after = ADES_CTRY)
+colSums(is.na(base_dataset))
+# No NA's - yaaayy!!
+glimpse(base_dataset)
+summary(base_dataset)
+
+```
+
+```{r}
+#quick peek
+base_dataset
+#
+#
+#
+#NOW IT'S TIME TO LOOP TO OTHER MONTHS AND BIND THEM ALL!
+#
+#
+#
+
+#quick first look
+temp1 <- base_dataset %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n()) %>% mutate(ROUTE = paste(ADEP_REG, ADES_REG, sep = "-"), .keep = "unused")
+
+temp1 %>% ggplot(aes(x = DATE)) +
+  geom_line(aes(y = (FLIGHTS), color = ROUTE))
 ```

### 2021-06-21 paper.Rmd
please loop

@@
 ```{r}
+###Preparatory codes
+
 #If someone needs filters below, here we can control any selective sample
@@
 #Currently, those filters are not in use in the code, they are here just in case.
-
 bra_10_apts <- c("SBBR", "SBGR", "SBSP", "SBKP", "SBRJ", "SBGL", "SBCF", "SBSV", "SBPA", "SBCT")
@@
 study_airports <- c(bra_10_apts, eur_apts)
+
+###Preparing airport file
+#NOTE: the airports.csv file below is the downloaded file from www.ourairports.org.
+airports <- read_csv("data-raw/airports.csv") %>% mutate(continent = case_when(is.na(continent) ~ "NA", TRUE ~ continent))
+
+#There are missing airports and too much variables
+apt_countries <- airports %>% transmute(ICAO = ident, CTRY = iso_country) %>% add_row(ICAO  = c("SPJC", "YBMC", "LSZM", "YSCH", "EPLB", "K3M3", "VV01", "SC28", "CWF2", "EHDB", "74XA", "HE13"), CTRY  = c("PE", "AU", "FR", "AU", "PL", "US", "VN", "US", "CA", "NL", "US", "EG"))
+
+#If you need to write
+#write_csv(apt_countries, "./data/apt_countries.csv")
+#apt_countries is ready.
+
+# Associate the regions
+a <- airports %>% filter(continent == "EU") %>% select(iso_country) %>% unique()
+eur_countries <- a$iso_country
+#eur_countries is ready.
+
 ```
@@
 1.2.1 Gather and prepare data
+
 ```{r}
@@
 ```{r}
-#Joining countries and regions
-# NOTE: the airports.csv file below is the downloaded file from www.ourairports.org.
-
-airports <- read_csv("data-raw/airports.csv") %>% mutate(continent = case_when(is.na(continent) ~ "NA", TRUE ~ continent))
-apt_countries <- airports %>% transmute(ICAO = ident, CTRY = iso_country)
-
-#There are missing airports
-apt_countries <- apt_countries %>% add_row(ICAO  = c("SPJC", "YBMC", "LSZM", "YSCH", "EPLB", "K3M3", "VV01", "SC28", "CWF2", "EHDB", "74XA", "HE13"), CTRY  = c("PE", "AU", "FR", "AU", "PL", "US", "VN", "US", "CA", "NL", "US", "EG"))
-
-#If you need to write
-#write_csv(apt_countries, "./data/apt_countries.csv")
-
 #Joining to ADEP
@@
 fb4 <- fb3 %>% drop_na() %>% mutate(ADEP_CTRY = as.factor(ADEP_CTRY), ADES_CTRY = as.factor(ADES_CTRY))
-
-# Associate the regions
-a <- airports %>% filter(continent == "EU") %>% select(iso_country) %>% unique()
-eur_countries <- a$iso_country
 

### 2021-06-21 R/please_loop.R

@@ -1,0 +1,24 @@
+open_sky <- read.csv("./data-raw/flightlist_20190101_20190131.csv") %>% transmute(ADEP = origin, ADES = destination, TYPE = as.factor(typecode), DATE = date(day), CALL = callsign, ACFT_ID = aircraft_uid)
+
+#Easily "dropping NA's# - this can be further sofisticated
+fb <- open_sky %>% drop_na()
+
+#Joining to ADEP
+fb2 <- left_join(fb, apt_countries, by = c("ADEP" = "ICAO")) %>% mutate(ADEP_CTRY = CTRY, .keep = "unused", .after = ADEP)
+#Joining to ADES
+fb3 <- left_join(fb2, apt_countries, by = c("ADES" = "ICAO")) %>% mutate(ADES_CTRY = CTRY, .keep = "unused", .after = ADES)
+
+fb4 <- fb3 %>% drop_na() %>% mutate(ADEP_CTRY = as.factor(ADEP_CTRY), ADES_CTRY = as.factor(ADES_CTRY))
+
+# Associate the regions
+
+base_dataset <- fb4 %>%
+  mutate(ADEP_REG = as.factor(case_when(ADEP_CTRY == "US" ~ "US",
+                                                              ADEP_CTRY == "BR" ~ "BR",
+                                                              ADEP_CTRY %in% eur_countries ~ "EU",
+                                                              TRUE ~ "Other")), .after = ADEP_CTRY) %>%
+  mutate(ADES_REG = as.factor(case_when(ADES_CTRY == "US" ~ "US",
+                                        ADES_CTRY == "BR" ~ "BR",
+                                        ADES_CTRY %in% eur_countries ~ "EU",
+                                        TRUE ~ "Other")), .after = ADES_CTRY)
+rm(fb,fb2, fb3, fb4)

### 2021-06-21 paper.Rmd
first stab into different levels

@@
 library(lubridate)
-
+library(stats)
+library(zoo)
 ```
@@
 #Here I will assign only one month - "202105". If you want to include a full year, just assign year to "2021" or "2020". It works.
-year <- "202105"
+year <- "2021"
 file_names <- list_apt_files(.year = year)
@@
 #
+```
 
+```{r}
 #quick first look
-temp1 <- base_dataset %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n()) %>% mutate(ROUTE = paste(ADEP_REG, ADES_REG, sep = "-"), .keep = "unused")
+temp1 <- base_dataset %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n(), .groups = "drop") %>%
+  filter(ADEP_REG %in% c("BR", "EU"), ADES_REG  %in% c("BR", "EU")) %>%
+  mutate(ROUTE = paste(ADEP_REG, ADES_REG, sep = "-"), .keep = "unused", .before = "FLIGHTS") %>%
+  pivot_wider(names_from = ROUTE, values_from = FLIGHTS) 
+temp1
 
+#First stab at different levels of traffic
+
+n <- 0.5
 temp1 %>% ggplot(aes(x = DATE)) +
-  geom_line(aes(y = (FLIGHTS), color = ROUTE))
+  geom_point(aes(y = `EU-EU`, color = `EU-EU` > quantile(`EU-EU`[month(DATE) %in% 1:2], probs = n)), shape = 4) +
+  geom_point(aes(y = `BR-BR`, color = `BR-BR` > quantile(`BR-BR`[month(DATE) %in% 1:2], probs = n)), shape = 1) +
+  labs(y  = "Flights") + 
+  theme(legend.position = "bottom")
+
+#Normalized by the median of the last 3 months
+temp2 <- temp1 %>% pivot_longer(cols = 2:5, names_to = "ROUTE", values_to = "FLIGHTS") %>% group_by(ROUTE) %>% mutate(MOVING_MEDIAN = quantile(FLIGHTS[month(DATE) %in% month(last(DATE))-1:month(last(DATE))], probs = 0.5), NORM_FLTS = FLIGHTS/MOVING_MEDIAN)
+temp2
+
+temp2 %>% ggplot(aes(x = DATE)) +
+  geom_line(aes(y = NORM_FLTS, color = ROUTE)) +
+  theme_minimal()
+```
+
+```{r}
+
 ```

### 2021-06-21 paper.Rmd
minor correction

@@
 
-#Normalized by the median of the last 3 months
-temp2 <- temp1 %>% pivot_longer(cols = 2:5, names_to = "ROUTE", values_to = "FLIGHTS") %>% group_by(ROUTE) %>% mutate(MOVING_MEDIAN = quantile(FLIGHTS[month(DATE) %in% month(last(DATE))-1:month(last(DATE))], probs = 0.5), NORM_FLTS = FLIGHTS/MOVING_MEDIAN)
+#Normalized by the median of the last 3 months of the dataset
+temp2 <- temp1 %>% pivot_longer(cols = 2:5, names_to = "ROUTE", values_to = "FLIGHTS") %>% group_by(ROUTE) %>% mutate(MOVING_MEDIAN = quantile(FLIGHTS[month(DATE) %in% month(last(DATE))-2:month(last(DATE))], probs = 0.5), NORM_FLTS = FLIGHTS/MOVING_MEDIAN)
 temp2

### 2021-06-21 x0_data-preparation.Rmd
data prep Rmd

---
title: "Data Preparation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)

#Reading raw data
## todo: change utility function list_osn_files
source("./R/list_apt_files.R")
```

# Overview

This file serves as a jump pad for the data preparation (and will be updated as we go).
The utility functions introduced read in the data, perform some associations (e.g. country assignment), filter, and save out the flight-by-flight dataset to drive the study for the DASC paper.

Conventions:

* Opensky Network data downloaded from https://zenodo.org/record/4893103#.YNCqukxCSUk
* airports.csv downloaded from www.ourairports.org.
* downloaded data stored in sub-folder **data-raw**
* trimmed/cleaned data saved out to **data**

More to come. tbd!

# Read in OSN Data

```{r}
###Preparatory codes

#If someone needs filters below, here we can control any selective sample
#Filtering Brazilian and European data to reduce the sample (and csv) size
#Currently, those filters are not in use in the code, they are here just in case.
bra_10_apts <- c("SBBR", "SBGR", "SBSP", "SBKP", "SBRJ", "SBGL", "SBCF", "SBSV", "SBPA", "SBCT")
eur_apts <- c("EHAM","LFPG","EGLL","EDDF","EDDM","LEMD","LIRF","LEBL","EGKK","LSZH")
study_airports <- c(bra_10_apts, eur_apts)

###Preparing airport file
#NOTE: the airports.csv file below is the downloaded file from www.ourairports.org.
airports <- read_csv("data-raw/airports.csv") %>% mutate(continent = case_when(is.na(continent) ~ "NA", TRUE ~ continent))

#There are missing airports and too much variables
apt_countries <- airports %>% transmute(ICAO = ident, CTRY = iso_country) %>% add_row(ICAO  = c("SPJC", "YBMC", "LSZM", "YSCH", "EPLB", "K3M3", "VV01", "SC28", "CWF2", "EHDB", "74XA", "HE13"), CTRY  = c("PE", "AU", "FR", "AU", "PL", "US", "VN", "US", "CA", "NL", "US", "EG"))

#If you need to write
#write_csv(apt_countries, "./data/apt_countries.csv")
#apt_countries is ready.

# Associate the regions
a <- airports %>% filter(continent == "EU") %>% select(iso_country) %>% unique()
eur_countries <- a$iso_country
#eur_countries is ready.

```

## Reading the raw data

### Utility function for reading OSN and saving out analytic data

```{r}
#Here I will assign only one month - "202105". If you want to include a full year, just assign year to "2021" or "2020". It works.
#year <- "202105"
# file_names <- list_apt_files(.year = year)
#open_sky <- map_dfr(file_names, read_csv)

# column specification for OSN -----------------------------------
osn_colspec <- cols(
  callsign = col_character(),
  number = col_character(),
  icao24 = col_character(),
  registration = col_character(),
  typecode = col_character(),
  origin = col_character(),
  destination = col_character(),
  firstseen = col_datetime(format = ""),
  lastseen = col_datetime(format = ""),
  day = col_datetime(format = ""),
  latitude_1 = col_double(),
  longitude_1 = col_double(),
  altitude_1 = col_double(),
  latitude_2 = col_double(),
  longitude_2 = col_double(),
  altitude_2 = col_double()
)

#' read_osn_covid
#' 
#' utility function to read one month of OSN covid data
#'
#' @param .fn filename of OSN data file
#' @param .col_spec dplyr column specification for datatype assignment
#'
read_osn_covid <- function(.fn, .col_spec = osn_colspec){
# todo - check filename

# read 
  fb <- read_csv(file = .fn, col_types = .col_spec)

# Selecting relevant variables

  fb <- fb %>% 
  transmute( ADEP = origin
            ,ADES = destination
            ,TYPE = as.factor(typecode)
            ,DATE = date(day)
            ,CALL = callsign
            #,ACFT_ID = aircraft_uid    # throws an error in today's data download
            )

return(fb)
}

#-------------- add ADEP/ADES & country
#' append country code to ADEP and ADES
#'
#' @param .ds flight data set with ADEP and ADES
#' @param .apt_countries lookup of ADEP/ADES and country
#'
#' @return appended dataset
#' @export
#'
append_ad_ctry <- function(.ds, .apt_countries = apt_countries){
  # todo - check for valid dataset
  df <- .ds
  #Joining to ADEP
  df <- left_join(df, .apt_countries, by = c("ADEP" = "ICAO")) %>%
    mutate(ADEP_CTRY = CTRY, .keep = "unused", .after = ADEP)
  #Joining to ADES
  df <- left_join(df, .apt_countries, by = c("ADES" = "ICAO")) %>% 
    mutate(ADES_CTRY = CTRY, .keep = "unused", .after = ADES)
  return(df)
}

#------------------ add region
#' Add region to identified ADEP/ADES country
#'
#' @param .ds flight data set with ADEP_CTRY and ADES_CTRY
#'
#' @return appended data set
#' @export
#'
append_region <- function(.ds, .eur_countries = eur_countries){
  # todo check for valid data set
  df <- .ds
  # append region for study
  df <- df %>%
    # ADEP region
    mutate(ADEP_REG = case_when(
                          ADEP_CTRY == "US" ~ "US"
                         ,ADEP_CTRY == "BR" ~ "BR"
                         ,ADEP_CTRY %in% .eur_countries ~ "EU"
                         ,TRUE ~ "Other")
      ,.after = ADEP_CTRY
      ) %>%
    # ADES region
    mutate(ADES_REG = case_when(
                          ADES_CTRY == "US" ~ "US"
                         ,ADES_CTRY == "BR" ~ "BR"
                         ,ADES_CTRY %in% eur_countries ~ "EU"
                         ,TRUE ~ "Other")
      , .after = ADES_CTRY)
  return(df)
}

#---------------- check / describe data set
#' extract meta data for data set
#'
#' @param .ds covid flight data set
#'
#' @return meta dataset
#' @export
#'
extract_meta_osn <- function(.ds){
  # todo: check for valid data set
  df <- .ds 
  # extract useful meta-data
  df <- df %>% 
    mutate(MOF = floor_date(DATE, unit = "month")) %>%
    group_by(MOF) %>%
    summarise( N = n()                         # number of flights in data set
              ,ADEP_NA = sum(is.na(ADEP))      # number of ADEP NA
              ,ADES_NA = sum(is.na(ADES))
              ,TYPE_NA = sum(is.na(TYPE))
              )

  return(df)
}

check_and_write_meta <- function(.new_meta, .my_meta = "./data/meta_osn.csv"){
  our_meta <- .my_meta
  # meta file exists
  if(file.exists(our_meta)){
    meta <- read_csv(our_meta, col_types = cols())  # cöl_types = cols() to silence
    meta <- meta %>% bind_rows(.new_meta)
    # todo -------------- handle multiple run and changes
  }else{
    meta <- .new_meta
  }
  # store meta data
  write_csv(meta, file = our_meta)
}

prep_base_data <- function(.fn = file_names, ...){
ds <- read_osn_covid(.fn)
ds <- ds %>% append_ad_ctry()
ds <- ds %>% append_region()

meta <- extract_meta_osn(ds)
check_and_write_meta(meta)

#------------ daily counts -------------------------------------
daily_count <- ds %>%
  group_by(DATE) %>% summarise(FLTS = n())
check_and_write_meta(daily_count, "./data/daily_osn.csv")
#----------------------------------------------------------------

ds <- ds %>% drop_na()  # drop all ~ assumed to be clean

# write out base data
my_fn <- gsub(pattern = "-raw", replacement = "", x = .fn)
my_fn <- gsub(pattern = "flightlist_", replacement = "base_osn_", x = my_fn)
my_fn <- gsub(pattern = "\\.gz", replacement = "", x = my_fn)
write_csv(ds, file = my_fn)
message(paste0("run completed for ", my_fn))
}
```

## loop over all data

```{r}
# read all files (note adpat function name)
fns <- list_apt_files()

# crunch study data
walk(fns, .f = prep_base_data)
```

# Todo's, observations, etc.

The following is currently kept and not further processed.

## check for some observations

```{}
# Very few NA's - it's safe to drop and factor now
# todo: check what airports have no country association in ourairports.com
# fb2 %>% filter(is.na(ADEP_CTRY)) %>% group_by(ADEP) %>% count() %>% arrange(desc(n))
# A tibble: 11 x 2
# Groups:   ADEP [11]
#    ADEP      n
#    <chr> <int>
#  1 YPEC    128
#  2 EG32     88
#  3 EGZJ     31
#  4 YCPP     26
#  5 KMAN     16
#  6 KX40     15
#  7 CSV8      9
#  8 CARK      8
#  9 LHKK      2
# 10 CLV2      1
# 11 EKVO      1

fb4 <- fb3 %>% drop_na() %>% mutate(ADEP_CTRY = as.factor(ADEP_CTRY), ADES_CTRY = as.factor(ADES_CTRY))

#NOTE: Here we can adjust the European countries that comprises EU region by editing the "eur_countries" vector:
# Currently in european countries: "GB" "AD" "ES" "AL" "XK" "AT" "BA" "BE" "BG" "IS" "BY" "UA" "CH" "CZ" "SK" "DE" "RU" "DK" "HR" "EE" "FI" "GG" "JE" "IM" "NL" "IE" "FO" "LU" "NO" "PL" "PT" "SE" "LV" "LT" "FR" "GR" "HU" "IT" "LI" "SI" "MT" "MC" "RO" "TR" "MD" "MK" "GI" "RS" "ME" "SM" "GE" "VA"

base_dataset <- fb4 %>% mutate(ADEP_REG = as.factor(case_when(ADEP_CTRY == "US" ~ "US",
                                    ADEP_CTRY == "BR" ~ "BR",
                                    ADEP_CTRY %in% eur_countries ~ "EU",
                                    TRUE ~ "Other")), .after = ADEP_CTRY) %>%
                mutate(ADES_REG = as.factor(case_when(ADES_CTRY == "US" ~ "US",
                                    ADES_CTRY == "BR" ~ "BR",
                                    ADES_CTRY %in% eur_countries ~ "EU",
                                    TRUE ~ "Other")), .after = ADES_CTRY)
colSums(is.na(base_dataset))
# No NA's - yaaayy!!
glimpse(base_dataset)
summary(base_dataset)

```

```{r, eval=FALSE}
#quick peek
base_dataset
#
#
#
#NOW IT'S TIME TO LOOP TO OTHER MONTHS AND BIND THEM ALL!
#
#
#

#quick first look
temp1 <- base_dataset %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n()) %>% mutate(ROUTE = paste(ADEP_REG, ADES_REG, sep = "-"), .keep = "unused")

temp1 %>% ggplot(aes(x = DATE)) +
  geom_line(aes(y = (FLIGHTS), color = ROUTE))
```

### 2021-06-21 paper.Rmd
initial hookup with daily flights

@@
 
-```{r}
+```{r globalstuff, message=FALSE}
+## set bookdown specs/defaults for high quality output
+#---------- check for the settings
+
+## theme default for ggplot
+theme_set(
+  theme_minimal()
+  )
+
 ###Preparatory codes
@@
 
+This study builds on publicly available data.
+Opensky Network collects crowdsourced air traffic data from more than 2500 feeders (sensor stations).
+To support the process of illustrating and studying the impact of the COVID pandemic on air traffic demand, a flight-by-flight dataset is provided on a monthly basis [@xavier_olive_2021_4893103].
+The data spans the period since 1.
+January 2019.
+Fig.
+\@ref(fig:osndaily) shows the number of daily flights tracked by Opensky Network globally.
+
+<!-- add some data set description, e.g. number of files, etc -->
+
+(ref:osndaily) Daily flights tracked by Opensky Network
+
+```{r osndaily, fig.cap="(ref:osndaily)", message=FALSE}
+daily_tfc <- read_csv("./data/daily_osn.csv")
+
+daily_tfc %>%
+  ggplot(mapping = aes(x = DATE, y = FLTS)) +
+  geom_line() +
+  labs(x = NULL, y = "flights")
+```
+
 # Results/Discussion
@@
 
-fb <- open_sky %>% transmute(ADEP = origin, ADES = destination, TYPE = as.factor(typecode), DATE = date(day), CALL = callsign, ACFT_ID = aircraft_uid)
+fb <- open_sky %>% transmute(ADEP = origin, ADES = destination, TYPE = as.factor(typecode), DATE = date(day), CALL = callsign
+                             #, ACFT_ID = aircraft_uid
+                             )
 

### 2021-06-21 paper.tex
initial hookup with daily flights

@@
 \usepackage{longtable,booktabs}
+\usepackage{graphicx}
+% We will generate all images so they have a width \maxwidth. This means
+% that they will get their normal width if they fit onto the page, but
+% are scaled down if they would overflow the margins.
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
+\else\Gin@nat@width\fi}
+\makeatother
+\let\Oldincludegraphics\includegraphics
+\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
 
@@
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\# set bookdown specs/defaults for high quality output}
+\CommentTok{\#{-}{-}{-}{-}{-}{-}{-}{-}{-}{-} check for the settings}
+
+\DocumentationTok{\#\# theme default for ggplot}
+\FunctionTok{theme\_set}\NormalTok{(}
+  \FunctionTok{theme\_minimal}\NormalTok{()}
+\NormalTok{  )}
+
+\DocumentationTok{\#\#\#Preparatory codes}
+
+\CommentTok{\#If someone needs filters below, here we can control any selective sample}
+\CommentTok{\#Filtering Brazilian and European data to reduce the sample (and csv) size}
+\CommentTok{\#Currently, those filters are not in use in the code, they are here just in case.}
+\NormalTok{bra\_10\_apts }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"SBBR"}\NormalTok{, }\StringTok{"SBGR"}\NormalTok{, }\StringTok{"SBSP"}\NormalTok{, }\StringTok{"SBKP"}\NormalTok{, }\StringTok{"SBRJ"}\NormalTok{, }\StringTok{"SBGL"}\NormalTok{, }\StringTok{"SBCF"}\NormalTok{, }\StringTok{"SBSV"}\NormalTok{, }\StringTok{"SBPA"}\NormalTok{, }\StringTok{"SBCT"}\NormalTok{)}
+\NormalTok{eur\_apts }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"EHAM"}\NormalTok{,}\StringTok{"LFPG"}\NormalTok{,}\StringTok{"EGLL"}\NormalTok{,}\StringTok{"EDDF"}\NormalTok{,}\StringTok{"EDDM"}\NormalTok{,}\StringTok{"LEMD"}\NormalTok{,}\StringTok{"LIRF"}\NormalTok{,}\StringTok{"LEBL"}\NormalTok{,}\StringTok{"EGKK"}\NormalTok{,}\StringTok{"LSZH"}\NormalTok{)}
+\NormalTok{study\_airports }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(bra\_10\_apts, eur\_apts)}
+
+\DocumentationTok{\#\#\#Preparing airport file}
+\CommentTok{\#NOTE: the airports.csv file below is the downloaded file from www.ourairports.org.}
+\NormalTok{airports }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"data{-}raw/airports.csv"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{continent =} \FunctionTok{case\_when}\NormalTok{(}\FunctionTok{is.na}\NormalTok{(continent) }\SpecialCharTok{\textasciitilde{}} \StringTok{"NA"}\NormalTok{, }\ConstantTok{TRUE} \SpecialCharTok{\textasciitilde{}}\NormalTok{ continent))}
+
+\CommentTok{\#There are missing airports and too much variables}
+\NormalTok{apt\_countries }\OtherTok{\textless{}{-}}\NormalTok{ airports }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{transmute}\NormalTok{(}\AttributeTok{ICAO =}\NormalTok{ ident, }\AttributeTok{CTRY =}\NormalTok{ iso\_country) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{add\_row}\NormalTok{(}\AttributeTok{ICAO  =} \FunctionTok{c}\NormalTok{(}\StringTok{"SPJC"}\NormalTok{, }\StringTok{"YBMC"}\NormalTok{, }\StringTok{"LSZM"}\NormalTok{, }\StringTok{"YSCH"}\NormalTok{, }\StringTok{"EPLB"}\NormalTok{, }\StringTok{"K3M3"}\NormalTok{, }\StringTok{"VV01"}\NormalTok{, }\StringTok{"SC28"}\NormalTok{, }\StringTok{"CWF2"}\NormalTok{, }\StringTok{"EHDB"}\NormalTok{, }\StringTok{"74XA"}\NormalTok{, }\StringTok{"HE13"}\NormalTok{), }\AttributeTok{CTRY  =} \FunctionTok{c}\NormalTok{(}\StringTok{"PE"}\NormalTok{, }\StringTok{"AU"}\NormalTok{, }\StringTok{"FR"}\NormalTok{, }\StringTok{"AU"}\NormalTok{, }\StringTok{"PL"}\NormalTok{, }\StringTok{"US"}\NormalTok{, }\StringTok{"VN"}\NormalTok{, }\StringTok{"US"}\NormalTok{, }\StringTok{"CA"}\NormalTok{, }\StringTok{"NL"}\NormalTok{, }\StringTok{"US"}\NormalTok{, }\StringTok{"EG"}\NormalTok{))}
+
+\CommentTok{\#If you need to write}
+\CommentTok{\#write\_csv(apt\_countries, "./data/apt\_countries.csv")}
+\CommentTok{\#apt\_countries is ready.}
+
+\CommentTok{\# Associate the regions}
+\NormalTok{a }\OtherTok{\textless{}{-}}\NormalTok{ airports }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(continent }\SpecialCharTok{==} \StringTok{"EU"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{select}\NormalTok{(iso\_country) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{unique}\NormalTok{()}
+\NormalTok{eur\_countries }\OtherTok{\textless{}{-}}\NormalTok{ a}\SpecialCharTok{$}\NormalTok{iso\_country}
+\CommentTok{\#eur\_countries is ready.}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#CURRENT COMMENTS AND TO{-}DO\textquotesingle{}S}
+\CommentTok{\# I HAVE DOWNLOADED 3 FILES FOR NOW (APR/19, APR/2020, APR/2021), JUST TO START "TIDYING" AND EXPLORING.}
+\CommentTok{\#IT\textquotesingle{}S IN THE DATA{-}RAW FOLDER (NOT SHARED WITH GITHUB), AS ALWAYS.}
+\end{Highlighting}
+\end{Shaded}
+
 \hypertarget{introduction}{%
@@
 \CommentTok{\# scenarios of low traffic? }
+\CommentTok{\#}
+\CommentTok{\# ???Research Question: }
+\CommentTok{\# ???RQ1.What was the impact of the pandemic on ATM resilience?}
+\CommentTok{\# ???  RQ1.1 How resilience can be modeled in a low{-}demand scenario?}
+\CommentTok{\# ???  RQ1.2 How resilient were different ATM systems worldwide?}
 \end{Highlighting}
@@
 \item
-  \ldots{}
+  ???RQ1.1 = through a qualitative analysis of previous proposed models
+\item
+  ???RQ1.2 = through a quantitative analysis of open data
 \end{itemize}
@@
 
+A mixed-method approach, based on:
+
+\begin{enumerate}
+\def\labelenumi{\alph{enumi})}
+\tightlist
+\item
+  to answer RQ1.1, a qualitative analysis of previous models to develop acute low-demand as a disturbance
+\item
+  to answer RQ1.2, a quantitative analysis of open data, to observe (or not) different levels/stages of stress/recovery, which could indicate different ``more'' or ``less'' resilience to the disturbances
+\end{enumerate}
+
 \hypertarget{open-source-data}{%
 \subsection{Open-source Data}\label{open-source-data}}
+
+This study builds on publicly available data.
+Opensky Network collects crowdsourced air traffic data from more than 2500 feeders (sensor stations).
+To support the process of illustrating and studying the impact of the COVID pandemic on air traffic demand, a flight-by-flight dataset is provided on a monthly basis (Olive, Strohmeier, and Lübbe 2021).
+The data spans the period since 1.
+January 2019.
+Fig.
+\ref{fig:osndaily} shows the number of daily flights tracked by Opensky Network globally.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{daily\_tfc }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"./data/daily\_osn.csv"}\NormalTok{)}
+
+\NormalTok{daily\_tfc }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{ggplot}\NormalTok{(}\AttributeTok{mapping =} \FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ DATE, }\AttributeTok{y =}\NormalTok{ FLTS)) }\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{() }\SpecialCharTok{+}
+  \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \ConstantTok{NULL}\NormalTok{, }\AttributeTok{y =} \StringTok{"flights"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/osndaily-1.pdf}
+\caption{\label{fig:osndaily}Daily flights tracked by Opensky Network}
+\end{figure}
 
@@
 
-to be developed
+1.1
+
+\begin{enumerate}
+\def\labelenumi{\alph{enumi})}
+\tightlist
+\item
+  Resilience can be measured as a function of time - the smaller the relationship between time of stress and the time of recovery, more resilient a system is.
+\end{enumerate}
+
+1.2 how to use open data to ``see'' resilience?
+
+1.2.1 Gather and prepare data
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#Reading raw data}
+
+\FunctionTok{source}\NormalTok{(}\StringTok{"./R/list\_apt\_files.R"}\NormalTok{)}
+
+\CommentTok{\#Here I will assign only one month {-} "202105". If you want to include a full year, just assign year to "2021" or "2020". It works.}
+\NormalTok{year }\OtherTok{\textless{}{-}} \StringTok{"2021"}
+\NormalTok{file\_names }\OtherTok{\textless{}{-}} \FunctionTok{list\_apt\_files}\NormalTok{(}\AttributeTok{.year =}\NormalTok{ year)}
+\NormalTok{open\_sky }\OtherTok{\textless{}{-}} \FunctionTok{map\_dfr}\NormalTok{(file\_names, read\_csv)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## 
+## -- Column specification --------------------------------------------------------
+## cols(
+##   callsign = col_character(),
+##   number = col_character(),
+##   icao24 = col_character(),
+##   registration = col_character(),
+##   typecode = col_character(),
+##   origin = col_character(),
+##   destination = col_character(),
+##   firstseen = col_datetime(format = ""),
+##   lastseen = col_datetime(format = ""),
+##   day = col_datetime(format = ""),
+##   latitude_1 = col_double(),
+##   longitude_1 = col_double(),
+##   altitude_1 = col_double(),
+##   latitude_2 = col_double(),
+##   longitude_2 = col_double(),
+##   altitude_2 = col_double()
+## )
+## 
+## 
+## -- Column specification --------------------------------------------------------
+## cols(
+##   callsign = col_character(),
+##   number = col_character(),
+##   icao24 = col_character(),
+##   registration = col_character(),
+##   typecode = col_character(),
+##   origin = col_character(),
+##   destination = col_character(),
+##   firstseen = col_datetime(format = ""),
+##   lastseen = col_datetime(format = ""),
+##   day = col_datetime(format = ""),
+##   latitude_1 = col_double(),
+##   longitude_1 = col_double(),
+##   altitude_1 = col_double(),
+##   latitude_2 = col_double(),
+##   longitude_2 = col_double(),
+##   altitude_2 = col_double()
+## )
+## 
+## 
+## -- Column specification --------------------------------------------------------
+## cols(
+##   callsign = col_character(),
+##   number = col_character(),
+##   icao24 = col_character(),
+##   registration = col_character(),
+##   typecode = col_character(),
+##   origin = col_character(),
+##   destination = col_character(),
+##   firstseen = col_datetime(format = ""),
+##   lastseen = col_datetime(format = ""),
+##   day = col_datetime(format = ""),
+##   latitude_1 = col_double(),
+##   longitude_1 = col_double(),
+##   altitude_1 = col_double(),
+##   latitude_2 = col_double(),
+##   longitude_2 = col_double(),
+##   altitude_2 = col_double()
+## )
+## 
+## 
+## -- Column specification --------------------------------------------------------
+## cols(
+##   callsign = col_character(),
+##   number = col_character(),
+##   icao24 = col_character(),
+##   registration = col_character(),
+##   typecode = col_character(),
+##   origin = col_character(),
+##   destination = col_character(),
+##   firstseen = col_datetime(format = ""),
+##   lastseen = col_datetime(format = ""),
+##   day = col_datetime(format = ""),
+##   latitude_1 = col_double(),
+##   longitude_1 = col_double(),
+##   altitude_1 = col_double(),
+##   latitude_2 = col_double(),
+##   longitude_2 = col_double(),
+##   altitude_2 = col_double()
+## )
+## 
+## 
+## -- Column specification --------------------------------------------------------
+## cols(
+##   callsign = col_character(),
+##   number = col_character(),
+##   icao24 = col_character(),
+##   registration = col_character(),
+##   typecode = col_character(),
+##   origin = col_character(),
+##   destination = col_character(),
+##   firstseen = col_datetime(format = ""),
+##   lastseen = col_datetime(format = ""),
+##   day = col_datetime(format = ""),
+##   latitude_1 = col_double(),
+##   longitude_1 = col_double(),
+##   altitude_1 = col_double(),
+##   latitude_2 = col_double(),
+##   longitude_2 = col_double(),
+##   altitude_2 = col_double()
+## )
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#Selecting relevant variables}
+
+\NormalTok{fb }\OtherTok{\textless{}{-}}\NormalTok{ open\_sky }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{transmute}\NormalTok{(}\AttributeTok{ADEP =}\NormalTok{ origin, }\AttributeTok{ADES =}\NormalTok{ destination, }\AttributeTok{TYPE =} \FunctionTok{as.factor}\NormalTok{(typecode), }\AttributeTok{DATE =} \FunctionTok{date}\NormalTok{(day), }\AttributeTok{CALL =}\NormalTok{ callsign}
+                             \CommentTok{\#, ACFT\_ID = aircraft\_uid}
+\NormalTok{                             )}
+
+\CommentTok{\#Easily "dropping NA\textquotesingle{}s\# {-} this can be further sofisticated}
+
+\NormalTok{fb1 }\OtherTok{\textless{}{-}}\NormalTok{ fb }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{drop\_na}\NormalTok{()}
+\NormalTok{fb1}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## # A tibble: 3,338,649 x 5
+##    ADEP  ADES  TYPE  DATE       CALL   
+##    <chr> <chr> <fct> <date>     <chr>  
+##  1 KJFK  LSGG  B788  2021-01-01 ETH726 
+##  2 KMIA  KMIA  B763  2021-01-01 LCO1108
+##  3 VHHH  71KY  G650  2021-01-01 ABW9515
+##  4 OMDM  YSSY  A343  2021-01-01 ASY052 
+##  5 KLAX  EDDF  B77L  2021-01-01 CSN461 
+##  6 EBLG  EBLG  B744  2021-01-01 ATG6652
+##  7 KORD  EHAM  B77L  2021-01-01 CSN5203
+##  8 FAOR  OMDB  B738  2021-01-01 KQA304 
+##  9 YSSY  OMDB  B77W  2021-01-01 UAE415 
+## 10 YSSY  WSSS  A332  2021-01-01 ALK302 
+## # ... with 3,338,639 more rows
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#Joining to ADEP}
+\NormalTok{fb2 }\OtherTok{\textless{}{-}} \FunctionTok{left\_join}\NormalTok{(fb1, apt\_countries, }\AttributeTok{by =} \FunctionTok{c}\NormalTok{(}\StringTok{"ADEP"} \OtherTok{=} \StringTok{"ICAO"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{ADEP\_CTRY =}\NormalTok{ CTRY, }\AttributeTok{.keep =} \StringTok{"unused"}\NormalTok{, }\AttributeTok{.after =}\NormalTok{ ADEP)}
+\CommentTok{\#Joining to ADES}
+\NormalTok{fb3 }\OtherTok{\textless{}{-}} \FunctionTok{left\_join}\NormalTok{(fb2, apt\_countries, }\AttributeTok{by =} \FunctionTok{c}\NormalTok{(}\StringTok{"ADES"} \OtherTok{=} \StringTok{"ICAO"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{ADES\_CTRY =}\NormalTok{ CTRY, }\AttributeTok{.keep =} \StringTok{"unused"}\NormalTok{, }\AttributeTok{.after =}\NormalTok{ ADES)}
+
+\CommentTok{\#Check  NA\textquotesingle{}s}
+\FunctionTok{colSums}\NormalTok{(}\FunctionTok{is.na}\NormalTok{(fb3))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##      ADEP ADEP_CTRY      ADES ADES_CTRY      TYPE      DATE      CALL 
+##         0      1271         0      1446         0         0         0
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# Very few NA\textquotesingle{}s {-} it\textquotesingle{}s safe to drop and factor now}
+
+\NormalTok{fb4 }\OtherTok{\textless{}{-}}\NormalTok{ fb3 }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{drop\_na}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{ADEP\_CTRY =} \FunctionTok{as.factor}\NormalTok{(ADEP\_CTRY), }\AttributeTok{ADES\_CTRY =} \FunctionTok{as.factor}\NormalTok{(ADES\_CTRY))}
+
+\CommentTok{\#NOTE: Here we can adjust the European countries that comprises EU region by editing the "eur\_countries" vector:}
+\CommentTok{\# Currently in european countries: "GB" "AD" "ES" "AL" "XK" "AT" "BA" "BE" "BG" "IS" "BY" "UA" "CH" "CZ" "SK" "DE" "RU" "DK" "HR" "EE" "FI" "GG" "JE" "IM" "NL" "IE" "FO" "LU" "NO" "PL" "PT" "SE" "LV" "LT" "FR" "GR" "HU" "IT" "LI" "SI" "MT" "MC" "RO" "TR" "MD" "MK" "GI" "RS" "ME" "SM" "GE" "VA"}
+
+\NormalTok{base\_dataset }\OtherTok{\textless{}{-}}\NormalTok{ fb4 }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{ADEP\_REG =} \FunctionTok{as.factor}\NormalTok{(}\FunctionTok{case\_when}\NormalTok{(ADEP\_CTRY }\SpecialCharTok{==} \StringTok{"US"} \SpecialCharTok{\textasciitilde{}} \StringTok{"US"}\NormalTok{,}
+\NormalTok{                                    ADEP\_CTRY }\SpecialCharTok{==} \StringTok{"BR"} \SpecialCharTok{\textasciitilde{}} \StringTok{"BR"}\NormalTok{,}
+\NormalTok{                                    ADEP\_CTRY }\SpecialCharTok{\%in\%}\NormalTok{ eur\_countries }\SpecialCharTok{\textasciitilde{}} \StringTok{"EU"}\NormalTok{,}
+                                    \ConstantTok{TRUE} \SpecialCharTok{\textasciitilde{}} \StringTok{"Other"}\NormalTok{)), }\AttributeTok{.after =}\NormalTok{ ADEP\_CTRY) }\SpecialCharTok{\%\textgreater{}\%}
+                \FunctionTok{mutate}\NormalTok{(}\AttributeTok{ADES\_REG =} \FunctionTok{as.factor}\NormalTok{(}\FunctionTok{case\_when}\NormalTok{(ADES\_CTRY }\SpecialCharTok{==} \StringTok{"US"} \SpecialCharTok{\textasciitilde{}} \StringTok{"US"}\NormalTok{,}
+\NormalTok{                                    ADES\_CTRY }\SpecialCharTok{==} \StringTok{"BR"} \SpecialCharTok{\textasciitilde{}} \StringTok{"BR"}\NormalTok{,}
+\NormalTok{                                    ADES\_CTRY }\SpecialCharTok{\%in\%}\NormalTok{ eur\_countries }\SpecialCharTok{\textasciitilde{}} \StringTok{"EU"}\NormalTok{,}
+                                    \ConstantTok{TRUE} \SpecialCharTok{\textasciitilde{}} \StringTok{"Other"}\NormalTok{)), }\AttributeTok{.after =}\NormalTok{ ADES\_CTRY)}
+\FunctionTok{colSums}\NormalTok{(}\FunctionTok{is.na}\NormalTok{(base\_dataset))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##      ADEP ADEP_CTRY  ADEP_REG      ADES ADES_CTRY  ADES_REG      TYPE      DATE 
+##         0         0         0         0         0         0         0         0 
+##      CALL 
+##         0
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# No NA\textquotesingle{}s {-} yaaayy!!}
+\FunctionTok{glimpse}\NormalTok{(base\_dataset)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Rows: 3,336,679
+## Columns: 9
+## $ ADEP      <chr> "KJFK", "KMIA", "VHHH", "OMDM", "KLAX", "EBLG", "KORD", "FAO~
+## $ ADEP_CTRY <fct> US, US, HK, AE, US, BE, US, ZA, AU, AU, TW, NL, TW, US, QA, ~
+## $ ADEP_REG  <fct> US, US, Other, Other, US, EU, US, Other, Other, Other, Other~
+## $ ADES      <chr> "LSGG", "KMIA", "71KY", "YSSY", "EDDF", "EBLG", "EHAM", "OMD~
+## $ ADES_CTRY <fct> CH, US, US, AU, DE, BE, NL, AE, AE, SG, US, PE, CA, AU, QA, ~
+## $ ADES_REG  <fct> EU, US, US, Other, EU, EU, EU, Other, Other, Other, US, Othe~
+## $ TYPE      <fct> B788, B763, G650, A343, B77L, B744, B77L, B738, B77W, A332, ~
+## $ DATE      <date> 2021-01-01, 2021-01-01, 2021-01-01, 2021-01-01, 2021-01-01,~
+## $ CALL      <chr> "ETH726", "LCO1108", "ABW9515", "ASY052", "CSN461", "ATG6652~
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{summary}\NormalTok{(base\_dataset)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##      ADEP             ADEP_CTRY        ADEP_REG           ADES          
+##  Length:3336679     US     :2225791   BR   :  35791   Length:3336679    
+##  Class :character   AU     : 127008   EU   : 596507   Class :character  
+##  Mode  :character   DE     : 109611   Other: 478590   Mode  :character  
+##                     GB     :  84635   US   :2225791                     
+##                     FR     :  64874                                     
+##                     CA     :  45627                                     
+##                     (Other): 679133                                     
+##    ADES_CTRY        ADES_REG            TYPE              DATE           
+##  US     :2223845   BR   :  35937   B738   : 356779   Min.   :2021-01-01  
+##  AU     : 127206   EU   : 597541   A320   : 234165   1st Qu.:2021-02-14  
+##  DE     : 109350   Other: 479356   B737   : 205056   Median :2021-03-26  
+##  GB     :  84605   US   :2223845   A321   : 135527   Mean   :2021-03-22  
+##  FR     :  65219                   A319   : 118359   3rd Qu.:2021-04-29  
+##  CA     :  45727                   E75L   : 112965   Max.   :2021-05-30  
+##  (Other): 680727                   (Other):2173828                       
+##      CALL          
+##  Length:3336679    
+##  Class :character  
+##  Mode  :character  
+##                    
+##                    
+##                    
+## 
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#quick peek}
+\NormalTok{base\_dataset}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## # A tibble: 3,336,679 x 9
+##    ADEP  ADEP_CTRY ADEP_REG ADES  ADES_CTRY ADES_REG TYPE  DATE       CALL   
+##    <chr> <fct>     <fct>    <chr> <fct>     <fct>    <fct> <date>     <chr>  
+##  1 KJFK  US        US       LSGG  CH        EU       B788  2021-01-01 ETH726 
+##  2 KMIA  US        US       KMIA  US        US       B763  2021-01-01 LCO1108
+##  3 VHHH  HK        Other    71KY  US        US       G650  2021-01-01 ABW9515
+##  4 OMDM  AE        Other    YSSY  AU        Other    A343  2021-01-01 ASY052 
+##  5 KLAX  US        US       EDDF  DE        EU       B77L  2021-01-01 CSN461 
+##  6 EBLG  BE        EU       EBLG  BE        EU       B744  2021-01-01 ATG6652
+##  7 KORD  US        US       EHAM  NL        EU       B77L  2021-01-01 CSN5203
+##  8 FAOR  ZA        Other    OMDB  AE        Other    B738  2021-01-01 KQA304 
+##  9 YSSY  AU        Other    OMDB  AE        Other    B77W  2021-01-01 UAE415 
+## 10 YSSY  AU        Other    WSSS  SG        Other    A332  2021-01-01 ALK302 
+## # ... with 3,336,669 more rows
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#}
+\CommentTok{\#}
+\CommentTok{\#}
+\CommentTok{\#NOW IT\textquotesingle{}S TIME TO LOOP TO OTHER MONTHS AND BIND THEM ALL!}
+\CommentTok{\#}
+\CommentTok{\#}
+\CommentTok{\#}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#quick first look}
+\NormalTok{temp1 }\OtherTok{\textless{}{-}}\NormalTok{ base\_dataset }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{group\_by}\NormalTok{(DATE, ADEP\_REG, ADES\_REG) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{summarize}\NormalTok{(}\AttributeTok{FLIGHTS =} \FunctionTok{n}\NormalTok{(), }\AttributeTok{.groups =} \StringTok{"drop"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{filter}\NormalTok{(ADEP\_REG }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\StringTok{"BR"}\NormalTok{, }\StringTok{"EU"}\NormalTok{), ADES\_REG  }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\StringTok{"BR"}\NormalTok{, }\StringTok{"EU"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{ROUTE =} \FunctionTok{paste}\NormalTok{(ADEP\_REG, ADES\_REG, }\AttributeTok{sep =} \StringTok{"{-}"}\NormalTok{), }\AttributeTok{.keep =} \StringTok{"unused"}\NormalTok{, }\AttributeTok{.before =} \StringTok{"FLIGHTS"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{pivot\_wider}\NormalTok{(}\AttributeTok{names\_from =}\NormalTok{ ROUTE, }\AttributeTok{values\_from =}\NormalTok{ FLIGHTS) }
+\NormalTok{temp1}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## # A tibble: 149 x 5
+##    DATE       `BR-BR` `BR-EU` `EU-BR` `EU-EU`
+##    <date>       <int>   <int>   <int>   <int>
+##  1 2021-01-01     189       8       7    1792
+##  2 2021-01-02     265       7      10    3088
+##  3 2021-01-03     264      12       6    3721
+##  4 2021-01-04     273       6      11    3446
+##  5 2021-01-05     320       5      11    3068
+##  6 2021-01-06     262      10       7    2930
+##  7 2021-01-07     288       8      11    3088
+##  8 2021-01-08     266       3       1    3343
+##  9 2021-01-09     253       7      11    2695
+## 10 2021-01-10     233       7       7    3371
+## # ... with 139 more rows
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#First stab at different levels of traffic}
+
+\NormalTok{n }\OtherTok{\textless{}{-}} \FloatTok{0.5}
+\NormalTok{temp1 }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ DATE)) }\SpecialCharTok{+}
+  \FunctionTok{geom\_point}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{y =} \StringTok{\textasciigrave{}}\AttributeTok{EU{-}EU}\StringTok{\textasciigrave{}}\NormalTok{, }\AttributeTok{color =} \StringTok{\textasciigrave{}}\AttributeTok{EU{-}EU}\StringTok{\textasciigrave{}} \SpecialCharTok{\textgreater{}} \FunctionTok{quantile}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{EU{-}EU}\StringTok{\textasciigrave{}}\NormalTok{[}\FunctionTok{month}\NormalTok{(DATE) }\SpecialCharTok{\%in\%} \DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{], }\AttributeTok{probs =}\NormalTok{ n)), }\AttributeTok{shape =} \DecValTok{4}\NormalTok{) }\SpecialCharTok{+}
+  \FunctionTok{geom\_point}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{y =} \StringTok{\textasciigrave{}}\AttributeTok{BR{-}BR}\StringTok{\textasciigrave{}}\NormalTok{, }\AttributeTok{color =} \StringTok{\textasciigrave{}}\AttributeTok{BR{-}BR}\StringTok{\textasciigrave{}} \SpecialCharTok{\textgreater{}} \FunctionTok{quantile}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{BR{-}BR}\StringTok{\textasciigrave{}}\NormalTok{[}\FunctionTok{month}\NormalTok{(DATE) }\SpecialCharTok{\%in\%} \DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{], }\AttributeTok{probs =}\NormalTok{ n)), }\AttributeTok{shape =} \DecValTok{1}\NormalTok{) }\SpecialCharTok{+}
+  \FunctionTok{labs}\NormalTok{(}\AttributeTok{y  =} \StringTok{"Flights"}\NormalTok{) }\SpecialCharTok{+} 
+  \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{paper_files/figure-latex/unnamed-chunk-6-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#Normalized by the median of the last 3 months of the dataset}
+\NormalTok{temp2 }\OtherTok{\textless{}{-}}\NormalTok{ temp1 }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{pivot\_longer}\NormalTok{(}\AttributeTok{cols =} \DecValTok{2}\SpecialCharTok{:}\DecValTok{5}\NormalTok{, }\AttributeTok{names\_to =} \StringTok{"ROUTE"}\NormalTok{, }\AttributeTok{values\_to =} \StringTok{"FLIGHTS"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{group\_by}\NormalTok{(ROUTE) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{MOVING\_MEDIAN =} \FunctionTok{quantile}\NormalTok{(FLIGHTS[}\FunctionTok{month}\NormalTok{(DATE) }\SpecialCharTok{\%in\%} \FunctionTok{month}\NormalTok{(}\FunctionTok{last}\NormalTok{(DATE))}\SpecialCharTok{{-}}\DecValTok{2}\SpecialCharTok{:}\FunctionTok{month}\NormalTok{(}\FunctionTok{last}\NormalTok{(DATE))], }\AttributeTok{probs =} \FloatTok{0.5}\NormalTok{), }\AttributeTok{NORM\_FLTS =}\NormalTok{ FLIGHTS}\SpecialCharTok{/}\NormalTok{MOVING\_MEDIAN)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in month(DATE) %in% month(last(DATE)) - 2:month(last(DATE)): Länge des längeren Objektes
+##       ist kein Vielfaches der Länge des kürzeren Objektes
+
+## Warning in month(DATE) %in% month(last(DATE)) - 2:month(last(DATE)): Länge des längeren Objektes
+##       ist kein Vielfaches der Länge des kürzeren Objektes
+
+## Warning in month(DATE) %in% month(last(DATE)) - 2:month(last(DATE)): Länge des längeren Objektes
+##       ist kein Vielfaches der Länge des kürzeren Objektes
+
+## Warning in month(DATE) %in% month(last(DATE)) - 2:month(last(DATE)): Länge des längeren Objektes
+##       ist kein Vielfaches der Länge des kürzeren Objektes
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{temp2}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## # A tibble: 596 x 5
+## # Groups:   ROUTE [4]
+##    DATE       ROUTE FLIGHTS MOVING_MEDIAN NORM_FLTS
+##    <date>     <chr>   <int>         <dbl>     <dbl>
+##  1 2021-01-01 BR-BR     189          214.     0.881
+##  2 2021-01-01 BR-EU       8            7      1.14 
+##  3 2021-01-01 EU-BR       7            9      0.778
+##  4 2021-01-01 EU-EU    1792         3397      0.528
+##  5 2021-01-02 BR-BR     265          214.     1.24 
+##  6 2021-01-02 BR-EU       7            7      1    
+##  7 2021-01-02 EU-BR      10            9      1.11 
+##  8 2021-01-02 EU-EU    3088         3397      0.909
+##  9 2021-01-03 BR-BR     264          214.     1.23 
+## 10 2021-01-03 BR-EU      12            7      1.71 
+## # ... with 586 more rows
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{temp2 }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ DATE)) }\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{y =}\NormalTok{ NORM\_FLTS, }\AttributeTok{color =}\NormalTok{ ROUTE)) }\SpecialCharTok{+}
+  \FunctionTok{theme\_minimal}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{paper_files/figure-latex/unnamed-chunk-6-2.pdf}
+
+1.2.2
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-xavier_olive_2021_4893103}{}%
+Olive, Xavier, Martin Strohmeier, and Jannis Lübbe. 2021. {``{Crowdsourced air traffic data from The OpenSky Network 2020}.''} Zenodo. \url{https://doi.org/10.5281/zenodo.4893103}.
+
 \leavevmode\hypertarget{ref-strohmeier_crowdsourced_2021}{}%

### 2021-06-24 paper.Rmd
initial stab after loops

@@
 
-```{r}
+```{r eval=FALSE, include=FALSE}
 #Reading raw data
@@
 
-```{r}
+```{r, eval=FALSE, include=FALSE}
 #Joining to ADEP
@@
 #quick peek
-base_dataset
-#
-#
-#
-#NOW IT'S TIME TO LOOP TO OTHER MONTHS AND BIND THEM ALL!
-#
-#
-#
-```
+#Read "x0_data-preparation.Rmd" first, then
+full_dataset
 
-```{r}
-#quick first look
-temp1 <- base_dataset %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n(), .groups = "drop") %>%
+temp <- full_dataset %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n(), .groups = "drop") %>%
   filter(ADEP_REG %in% c("BR", "EU"), ADES_REG  %in% c("BR", "EU")) %>%
@@
   pivot_wider(names_from = ROUTE, values_from = FLIGHTS) 
+temp
+
+#First Look
+temp1 <- full_dataset %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n()) %>% mutate(ROUTE = paste(ADEP_REG, ADES_REG, sep = "-"), .keep = "unused")
 temp1
+
+temp1 %>% ggplot(aes(x = DATE)) +
+  geom_line(aes(y = (FLIGHTS), color = ROUTE))
 
@@
 #Normalized by the median of the last 3 months of the dataset
-temp2 <- temp1 %>% pivot_longer(cols = 2:5, names_to = "ROUTE", values_to = "FLIGHTS") %>% group_by(ROUTE) %>% mutate(MOVING_MEDIAN = quantile(FLIGHTS[month(DATE) %in% month(last(DATE))-2:month(last(DATE))], probs = 0.5), NORM_FLTS = FLIGHTS/MOVING_MEDIAN)
+temp1 %>% pivot_longer(cols = 2:5, names_to = "ROUTE", values_to = "FLIGHTS") %>% group_by(ROUTE) %>%
+  mutate(MOVING_MEDIAN = quantile(FLIGHTS[month(DATE) %in% month(last(DATE))-2:month(last(DATE))], probs = 0.5), NORM_FLTS = FLIGHTS/MOVING_MEDIAN)
 temp2

### 2021-06-24 x0_data-preparation.Rmd
initial stab after loops

@@
 
-```{r, eval=FALSE}
+```{r, eval=TRUE}
 #quick peek
-base_dataset
-#
-#
-#
-#NOW IT'S TIME TO LOOP TO OTHER MONTHS AND BIND THEM ALL!
-#
-#
-#
 
+base_filenames <- list.files("./data/", pattern = "base_osn_*")
+base_fullnames <- c(paste0("./data/", base_filenames))
+full_dataset <- map_dfr(base_fullnames, read_csv)
+
+```
+
+```{r}
 #quick first look
-temp1 <- base_dataset %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n()) %>% mutate(ROUTE = paste(ADEP_REG, ADES_REG, sep = "-"), .keep = "unused")
+temp1 <- full_dataset %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n()) %>% mutate(ROUTE = paste(ADEP_REG, ADES_REG, sep = "-"), .keep = "unused")
+temp1
 

### 2021-06-27 paper-old-failed-compile.Rmd
some small add on, had to convert to MS Word as pdf kept failing. Not sure what goes wrong.

--- 2021-06-24 paper.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
 library(tidyverse)
@@
 ## set bookdown specs/defaults for high quality output
-#---------- check for the settings
+knitr::opts_chunk$set(
+   echo = FALSE, message = FALSE, warning = FALSE    # messages
+  , dpi = 320               #, dev.args = list(type = "cairo") #
+  , dev = "ragg_png"        # works better than cairo
+  , fig.retina = 2      # resolution and device
+#  , fig.align = "center", fig.show = "hold"          # positioning of figure
+#  , fig.width = 6, fig.asp = 0.618
+#  , out.width = "85%"                                # outside dim of figure
+                      )
 
@@
   )
+
+## load study data - as long as not outsourced to data prep
+fns <- list.files(path = "./data/", pattern = "base_osn_", full.names = TRUE )
+col_spec <- cols(
+  ADEP = col_character(),
+  ADEP_CTRY = col_character(),
+  ADEP_REG = col_character(),
+  ADES = col_character(),
+  ADES_CTRY = col_character(),
+  ADES_REG = col_character(),
+  TYPE = col_character(),
+  DATE = col_date(format = ""),
+  CALL = col_character()
+)
+ds  <- fns %>% map_dfr(.f = ~ read_csv(.x))
+```
+
+```{r eval=FALSE}
+## CHECK WHAT CAN BE REMOVED FROM HERE AND IS NOW OUTSOURCED TO DATA PREP
 
@@
 January 2019.
-Fig.
-\@ref(fig:osndaily) shows the number of daily flights tracked by Opensky Network globally.
+Fig. \@ref(fig:osndaily) shows the number of daily flights tracked by Opensky Network globally.
 
@@
 #Read "x0_data-preparation.Rmd" first, then
-full_dataset
+ds
 
-temp <- full_dataset %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n(), .groups = "drop") %>%
+temp <- ds %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n(), .groups = "drop") %>%
   filter(ADEP_REG %in% c("BR", "EU"), ADES_REG  %in% c("BR", "EU")) %>%
@@
   pivot_wider(names_from = ROUTE, values_from = FLIGHTS) 
-temp
+head(temp)
 
 #First Look
-temp1 <- full_dataset %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n()) %>% mutate(ROUTE = paste(ADEP_REG, ADES_REG, sep = "-"), .keep = "unused")
-temp1
+temp1 <- ds %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n()) %>% mutate(ROUTE = paste(ADEP_REG, ADES_REG, sep = "-"), .keep = "unused")
+head(temp1)
 
@@
   geom_line(aes(y = (FLIGHTS), color = ROUTE))
+```
 
+<!-- NEEDED TO INAVTIVATE THE FOLLOWING - I COULD NOT RECONSTRUCT IT -->
+
+```{r eval=FALSE}
 #First stab at different levels of traffic
@@
   theme(legend.position = "bottom")
+```
 
+<!-- needed to deactivate the following:
+* threw error with column 5
+* could not reconcile which operation would have created the 5th col
+-->
+
+```{r, eval=FALSE}
 #Normalized by the median of the last 3 months of the dataset
-temp1 %>% pivot_longer(cols = 2:5, names_to = "ROUTE", values_to = "FLIGHTS") %>% group_by(ROUTE) %>%
+temp1 %>% pivot_longer(cols = 2:5, names_to = "ROUTE", values_to = "FLIGHTS") %>%   
+  group_by(ROUTE) %>%
   mutate(MOVING_MEDIAN = quantile(FLIGHTS[month(DATE) %in% month(last(DATE))-2:month(last(DATE))], probs = 0.5), NORM_FLTS = FLIGHTS/MOVING_MEDIAN)
-temp2
+
+head(temp2)
 
@@
 ```
-
-```{r}
-
-```
-
-1.2.2
 

### 2021-06-29 x0_data-preparation.Rmd
initial thoughts on thresholds

@@
 ```
+
+```{r}
+#THRESHOLD AT A FIXED QUANTILE PER YEAR
+ds1 <- full_dataset %>% group_by(YR = year(DATE), DATE) %>% summarize(N = n()) %>% mutate(THRESHOLD = quantile(N, probs = 0.7))
+ds1
+
+ds1 %>% ggplot(aes(x = DATE)) +
+  geom_line(aes(y = THRESHOLD)) +
+  geom_point(aes(y = N, color = N > THRESHOLD))
+
+#CHECK WEEKLY DISTRIBUTION
+ds2 <- ds1 %>% mutate(WEEK_DAY = wday(DATE, label = TRUE), W_END = WEEK_DAY %in% c("Sat", "Sun"))
+ds2
+
+ds2 %>% ggplot(aes(x = DATE)) +
+  geom_line(aes(y = N, color = W_END))
+
+ds2 %>% ggplot() +
+  geom_freqpoly(aes(x = N, color = W_END))
+```

### 2021-06-29 x0_data-preparation.Rmd
included thresholds graph and formulas

--- 2021-06-24 x0_data-preparation.Rmd
@@
 library(lubridate)
+library(zoo)
 
@@
 ```
+
+```{r, dpi=480}
+#THRESHOLD AT ROLLING QUANTILES PER YEAR
+#Please install the "roll" package.
+#install.packages("roll")
+library(roll)
+
+ds1 <- full_dataset %>% group_by(YR = year(DATE), DATE) %>% summarize(N = n(), .groups = "drop") %>% mutate(THRESHOLD = quantile(N, probs = 0.7))
+ds1
+
+ds1 %>% ggplot(aes(x = DATE)) +
+  geom_line(aes(y = THRESHOLD)) +
+  geom_point(aes(y = N, color = N > THRESHOLD))
+
+#CHECK WEEKLY DISTRIBUTION
+ds2 <- ds1 %>% mutate(WEEK_DAY = wday(DATE, label = TRUE), W_END = WEEK_DAY %in% c("Sat", "Sun"))
+uplim <- .95
+lowlim <- .20
+
+ds2 <- ds2 %>% mutate(MED = roll_quantile(N, width = 7, p = .50, min_obs = 7),
+                      UP_THR = roll_quantile(N, width = 180, p = uplim, min_obs = 7),
+                      LOW_THR = roll_quantile(N, width = 180, p = lowlim, min_obs = 7))
+ds2
+
+ds2 %>% ggplot(aes(x = DATE)) +
+  geom_col(aes(y = MED, fill = !(MED < LOW_THR | MED > UP_THR))) +
+  geom_line(aes(y = UP_THR), color = "blue") +
+  geom_line(aes(y = LOW_THR), color = "red") +
+  scale_x_date(limits = c(as.Date("2019-10-01"), as.Date("2020-12-01")), date_breaks = "30 days", date_labels = "%b\n%y") +
+  labs(y = "7-day rolling median", x = NULL, fill = "Normal operation?",
+       title = "What is normal?",
+       subtitle = "Red line - Lower threshold; Blue line - upper threshold",
+       caption = "Lower threshold set to 20th percentile of the last 180 days;\nUpper threshold set to 95th percentile of the last 180 days"
+)
+
+#ds2 %>% ggplot() +
+  geom_freqpoly(aes(x = N, color = W_END))
+```

### 2021-07-12 paper.Rmd
Covid data included

--- 2021-06-21 paper.Rmd
@@
 output:
-  bookdown::pdf_book:
-     base_format: rticles::ieee_article
+# failed to render and I could not finally track what kills the pdf
+#  bookdown::pdf_book:
+#    base_format: rticles::ieee_article
+  bookdown::word_document2
 bibliography: "DASC2021.bib"
@@
 library(zoo)
+
+# for corona data
+# install.packages("coronavirus"), but do not load it!
+# then run: coronavirus::update_dataset() 
 ```
 
-```{r globalstuff, message=FALSE}
+```{r, echo=FALSE, message=FALSE}
 ## set bookdown specs/defaults for high quality output
-#---------- check for the settings
+knitr::opts_chunk$set(
+   echo = FALSE, message = FALSE, warning = FALSE    # messages
+  , dpi = 320               #, dev.args = list(type = "cairo") #
+  , dev = "ragg_png"        # works better than cairo
+#  , fig.retina = 2         # !! throws error in reticles IEEE
+#  , fig.align = "center"   # !! throws error in reticles IEEE
+  , fig.show = "hold"          # positioning of figure
+  , fig.width = 6, fig.asp = 0.618
+#  , out.width = "95%"      # !! throws error in reticles IEEE
+)
 
@@
 
-###Preparatory codes
-
-#If someone needs filters below, here we can control any selective sample
-#Filtering Brazilian and European data to reduce the sample (and csv) size
-#Currently, those filters are not in use in the code, they are here just in case.
-bra_10_apts <- c("SBBR", "SBGR", "SBSP", "SBKP", "SBRJ", "SBGL", "SBCF", "SBSV", "SBPA", "SBCT")
-eur_apts <- c("EHAM","LFPG","EGLL","EDDF","EDDM","LEMD","LIRF","LEBL","EGKK","LSZH")
-study_airports <- c(bra_10_apts, eur_apts)
-
-###Preparing airport file
-#NOTE: the airports.csv file below is the downloaded file from www.ourairports.org.
-airports <- read_csv("data-raw/airports.csv") %>% mutate(continent = case_when(is.na(continent) ~ "NA", TRUE ~ continent))
-
-#There are missing airports and too much variables
-apt_countries <- airports %>% transmute(ICAO = ident, CTRY = iso_country) %>% add_row(ICAO  = c("SPJC", "YBMC", "LSZM", "YSCH", "EPLB", "K3M3", "VV01", "SC28", "CWF2", "EHDB", "74XA", "HE13"), CTRY  = c("PE", "AU", "FR", "AU", "PL", "US", "VN", "US", "CA", "NL", "US", "EG"))
-
-#If you need to write
-#write_csv(apt_countries, "./data/apt_countries.csv")
-#apt_countries is ready.
-
-# Associate the regions
-a <- airports %>% filter(continent == "EU") %>% select(iso_country) %>% unique()
-eur_countries <- a$iso_country
-#eur_countries is ready.
-
+## load study data - as long as not outsourced to data prep
+fns <- list.files(path = "./data/", pattern = "base_osn_", full.names = TRUE )
+col_spec <- cols(
+  ADEP = col_character(),
+  ADEP_CTRY = col_character(),
+  ADEP_REG = col_character(),
+  ADES = col_character(),
+  ADES_CTRY = col_character(),
+  ADES_REG = col_character(),
+  TYPE = col_character(),
+  DATE = col_date(format = ""),
+  CALL = col_character()
+)
+ds  <- fns %>% map_dfr(.f = ~ read_csv(.x))
 ```
@@
 ```{r}
-#CURRENT COMMENTS AND TO-DO'S
+# CURRENT COMMENTS AND TO-DO'S
 # I HAVE DOWNLOADED 3 FILES FOR NOW (APR/19, APR/2020, APR/2021), JUST TO START "TIDYING" AND EXPLORING.
-#IT'S IN THE DATA-RAW FOLDER (NOT SHARED WITH GITHUB), AS ALWAYS.
+# IT'S IN THE DATA-RAW FOLDER (NOT SHARED WITH GITHUB), AS ALWAYS.
 ```
@@
 
-However, the recent COVID-19 crisis posed a completely different, unexpected, and inverted challenge.
+On March 11, 2020, the World Health Organisation has declared the novel coronavirus (COVID-19) outbreak a global pandemic.
+The COVID-19 crisis posed a completely different, unexpected, and inverted challenge.
+Communicable disease control resulted in massive restrictions on international and regional air traffic and passenger travel. 
 Demand for air transport dropped as low as 90% of the previous "normal" in many places.
@@
 As a result, air navigation providers collected less fees for their services, due to significantly fewer flights.
-In the operational perspective, pilots and air traffic controllers practiced less.
-The problems and obstacles developed into many other dimensions.
-
-Hence, the current scenario is a proper moment to further investigate the concept of resilience.
+In the operational perspective, pilots and air traffic controllers practiced less. The problems and obstacles developed into many other dimensions.
 
@@
 # ???  RQ1.2 How resilient were different ATM systems worldwide?
-
 ```
@@
 <!--| 2. High-Level background of previous work or relevance. -->
+
+The unprecedented decline in air traffic demand resulted in severe financial strains on the air transport industry. 
+The financial support or lack thereof for airlines and airports has been widely covered in the media. 
+A variety of studies showed the interplay or consequences of the travel constraints. 
+To date, lower attention was given to the inherent change in terms of air transport services
+Hence, the current scenario is a proper moment to further investigate the concept of resilience.
 
@@
 
--   data-analytical approach - using open data / freely available (tbd: validated against organisational data)
--   ???RQ1.1 = through a qualitative analysis of previous proposed models
--   ???RQ1.2 = through a quantitative analysis of open data
+> data-analytical approach - using open data / freely available (tbd: validated against organisational data)
+
+This paper utilises a data-driven approach.
+Despite the fact that aviation is a data rich environment, operational data on the previous and current traffic levels is not consistently available. 
+The global traffic evaluation will be based on data from a community fed sensor network, Opensky-Network. 
+<<CHECK THIS>> The operational flight data will be enriched with other publicly available dataset supporting the description of the level of the COVID pandemic and national/regional responses to the pandemic.
+
+```{r} 
+# 
+#   ???RQ1.1 = through a qualitative analysis of previous proposed models
+#   ???RQ1.2 = through a quantitative analysis of open data
+```
 
@@
 
+  + novel methodology to describe disruption, transformation, and recovery phenomena of the air transport network <<stakeholder operations>>
+
 -   identification of patterns and/or measures to describe and quantify/evaluate the level of recovery (or disruption)
@@
 
+On 11. March 2020, the World Health Organisation reacted to the steadily increasing of infections and global spread of a newly detected Corona-Virus by declaring a pandemic.
+As an immediate response, many governments reacted by imposing controls to curb the further spread of communicable disease.
+Next to social distancing and recommended hygiene measures, this resulted in massive restrictions on international and regional air traffic and passenger travel.
+
+<!-- TODO: add diagram about increasing COVID infections/deaths and/or potential waves0 -->
+```{r}
+# coronavius (data) package is installed and updated
+# call it with coronavirus:: notation
+# for this paper, we reduce to May 2021 as update() will expand in future runs
+covid <- coronavirus::coronavirus %>% 
+  as_tibble() %>%
+  filter(date <= as.POSIXct("2021-06-01"))
+
+global_cov <- covid %>% 
+  group_by(type, date) %>%
+  summarise(total_cases = sum(cases), .groups = "drop") %>%
+  pivot_wider(names_from = type, values_from = total_cases) %>%
+  arrange(date) %>%
+  mutate(active = confirmed - death - recovered) %>%
+  mutate(active_total = cumsum(active),
+                recovered_total = cumsum(recovered),
+                death_total = cumsum(death))
+```
+
+(ref:fig-global-covid) Reported number of cases (22. Jan 2020 through 27. May 2021) [@coronavirus_data] based on data from John Hopkins University
+
+```{r fig-global-covid, fig.cap="(ref:fig-global-covid)"}
+tmp <- global_cov %>%
+  select(date, contains("total")) %>%
+  pivot_longer(cols = 2:4, names_to = "category", values_to = "counts")
+
+tmp %>%
+
+ggplot(aes(x = date, y = counts, fill = category)) +
+  geom_area() +
+  theme_minimal() +
+  labs(x = NULL, y = NULL) +
+  theme(legend.position = c(0.2, 0.3)) +
+  scale_fill_manual( name = "cases"
+                    ,values = c("orange","red", "green")
+                    ,labels = c("active", "death","recovered")) +
+  scale_y_continuous(labels = scales::label_number_si())
+```
+
 ## Resilience
+
+Resilience is a well-researched topic.
+The term is used by a diverse set of domains, communities, and research areas.
+In consequence there exists an abundance of theoretical definitions in the literature and numerous concepts that emerge from theses definitions and expert domains.
+
+The term originates from physics and is nowadays used with reference to safety, security, environment and ecological systems, mental health/psychology, biological system, and others.
+Despite the varying contexts, across all of these disciplines the concept of resilience is closely related with the capability and ability of the focus of concern - typically the system or agent - to return to a stable state after a disruption impacted the original state.
+
+Recently, the term is also used frequently on the political and strategic level.
+Policymakers, operational experts, and academes concur that the concept of resilience plays a major role when addressing and assessing the extent to which organisations and systems are prepared and capable to respond to and recover from disruptions.
+
+Within air navigation, the term 'resilience' has been picked up by several communities, both operational and scientific.
 
@@
 
+<!-- provide a pointer hook to the current interest in studying resilience (again) -->
+
+The unprecedented decline in air transportation since March 2020 has triggered an increased interest in the topic.
+For example under the umbrella of the ICAO Global Air Navigation Plan Study Group, an expert team is currently working on refining the performance framework with a view to resilience.
+With a view to COVID, there is a dual interest in understanding resilience in air navigation/transportation:
+
+-   political level: ADD SOME DESCRIPTION
+
+-   operational level: ADD SOME DESCRIPTION
+
+This paper addresses the operational dimension. Being able to characterise resilience within the operational domain will enable to address the more strategic and political decision-making.
+
 ## \<if we need to fill space\> Crowd-Sourced Data Collection
@@
 
+```{r osndata}
+daily_tfc <- read_csv("./data/daily_osn.csv") %>% distinct()
+
+osn_2019 <- daily_tfc %>% mutate(YR = lubridate::year(DATE)) %>%
+  filter(YR == 2019) %>%
+  summarise( DATE_MAX = DATE[FLTS == max(FLTS)], MAX = max(FLTS)
+            ,DATE_MIN = DATE[FLTS == min(FLTS)], MIN = min(FLTS))
+
+osn_2019_2 <- daily_tfc %>% 
+  filter(DATE >= lubridate::ymd("2019-01-01") & DATE < lubridate::ymd("2020-01-01")) %>%
+  summarise(MED = median(FLTS))
+```
+
 This study builds on publicly available data.
@@
 January 2019.
-Fig.
-\@ref(fig:osndaily) shows the number of daily flights tracked by Opensky Network globally.
-
-<!-- add some data set description, e.g. number of files, etc -->
+Fig. \@ref(fig:osndaily) shows the number of daily flights tracked by Opensky Network globally.
+The observed continual increase throughout the first six months in 2019 is driven by the increasing rate of ADSB equipage.
+Particularly, air transport operators in the United States or operating to the United States established compliance with the ADSB mandate applicable as of 1.
+January 2020.
+In 2019, the peak daily number of tracked flights ranges just under `r pretty(osn_2019$MAX, shrink = 0.1)[2]`.
+The negative spike observed on `r osn_2019$DATE_MIN` is linked to a data outage.
+Throughout the second half of 2019 the median number of flights ranged just under `r pretty(osn_2019_2$MED, shrink = 0.2)[2]`.
 
@@
 ```{r osndaily, fig.cap="(ref:osndaily)", message=FALSE}
-daily_tfc <- read_csv("./data/daily_osn.csv")
-
 daily_tfc %>%
@@
 ```
+
+<!-- add some data set description, e.g. number of files, etc -->
+
+## Measuring Resilience
+
+The concept of resilience (and robustness) is intimately linked with the construct of disruptions.
+
+(ref:resiliencebasics) Resilience as a function of disturbance impact
+
+```{r resiliencebasics, fig.cap="(ref:resiliencebasics)"}
+knitr::include_graphics("figures/resilience-timeline.png")
+```
+
+ Conceptually, resilience R can be measured as the observed loss in quality of service (performance) over the time to recovery, t1-t2, for a certain level of disruption.
+Thus, mathematically, this represents the area covered by
+
+$$\int_{a}^{b}[THR -LOS(t)]dt$$
+
+as presented in Fig. \@ref(fig:resiliencebasics) (LOS: loss of service / performance, THR: associated threshold).
 
@@
 
-```{r}
-#Reading raw data
+```{r, fig.cap="test"}
+#quick peek
+#Read "x0_data-preparation.Rmd" first, then
+head(ds)
 
-source("./R/list_apt_files.R")
-
-#Here I will assign only one month - "202105". If you want to include a full year, just assign year to "2021" or "2020". It works.
-year <- "2021"
-file_names <- list_apt_files(.year = year)
-open_sky <- map_dfr(file_names, read_csv)
-
-#Selecting relevant variables
-
-fb <- open_sky %>% transmute(ADEP = origin, ADES = destination, TYPE = as.factor(typecode), DATE = date(day), CALL = callsign
-                             #, ACFT_ID = aircraft_uid
-                             )
-
-#Easily "dropping NA's# - this can be further sofisticated
-
-fb1 <- fb %>% drop_na()
-fb1
-```
-
-```{r}
-#Joining to ADEP
-fb2 <- left_join(fb1, apt_countries, by = c("ADEP" = "ICAO")) %>% mutate(ADEP_CTRY = CTRY, .keep = "unused", .after = ADEP)
-#Joining to ADES
-fb3 <- left_join(fb2, apt_countries, by = c("ADES" = "ICAO")) %>% mutate(ADES_CTRY = CTRY, .keep = "unused", .after = ADES)
-
-#Check  NA's
-colSums(is.na(fb3))
-# Very few NA's - it's safe to drop and factor now
-
-fb4 <- fb3 %>% drop_na() %>% mutate(ADEP_CTRY = as.factor(ADEP_CTRY), ADES_CTRY = as.factor(ADES_CTRY))
-
-#NOTE: Here we can adjust the European countries that comprises EU region by editing the "eur_countries" vector:
-# Currently in european countries: "GB" "AD" "ES" "AL" "XK" "AT" "BA" "BE" "BG" "IS" "BY" "UA" "CH" "CZ" "SK" "DE" "RU" "DK" "HR" "EE" "FI" "GG" "JE" "IM" "NL" "IE" "FO" "LU" "NO" "PL" "PT" "SE" "LV" "LT" "FR" "GR" "HU" "IT" "LI" "SI" "MT" "MC" "RO" "TR" "MD" "MK" "GI" "RS" "ME" "SM" "GE" "VA"
-
-base_dataset <- fb4 %>% mutate(ADEP_REG = as.factor(case_when(ADEP_CTRY == "US" ~ "US",
-                                    ADEP_CTRY == "BR" ~ "BR",
-                                    ADEP_CTRY %in% eur_countries ~ "EU",
-                                    TRUE ~ "Other")), .after = ADEP_CTRY) %>%
-                mutate(ADES_REG = as.factor(case_when(ADES_CTRY == "US" ~ "US",
-                                    ADES_CTRY == "BR" ~ "BR",
-                                    ADES_CTRY %in% eur_countries ~ "EU",
-                                    TRUE ~ "Other")), .after = ADES_CTRY)
-colSums(is.na(base_dataset))
-# No NA's - yaaayy!!
-glimpse(base_dataset)
-summary(base_dataset)
-
-```
-
-```{r}
-#quick peek
-base_dataset
-#
-#
-#
-#NOW IT'S TIME TO LOOP TO OTHER MONTHS AND BIND THEM ALL!
-#
-#
-#
-```
-
-```{r}
-#quick first look
-temp1 <- base_dataset %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n(), .groups = "drop") %>%
+temp <- ds %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n(), .groups = "drop") %>%
   filter(ADEP_REG %in% c("BR", "EU"), ADES_REG  %in% c("BR", "EU")) %>%
@@
   pivot_wider(names_from = ROUTE, values_from = FLIGHTS) 
-temp1
+#head(temp)
 
-#First stab at different levels of traffic
+#First Look
+temp1 <- ds %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n()) %>% mutate(ROUTE = paste(ADEP_REG, ADES_REG, sep = "-"), .keep = "unused")
+#head(temp1)
 
-n <- 0.5
 temp1 %>% ggplot(aes(x = DATE)) +
-  geom_point(aes(y = `EU-EU`, color = `EU-EU` > quantile(`EU-EU`[month(DATE) %in% 1:2], probs = n)), shape = 4) +
-  geom_point(aes(y = `BR-BR`, color = `BR-BR` > quantile(`BR-BR`[month(DATE) %in% 1:2], probs = n)), shape = 1) +
-  labs(y  = "Flights") + 
-  theme(legend.position = "bottom")
-
-#Normalized by the median of the last 3 months of the dataset
-temp2 <- temp1 %>% pivot_longer(cols = 2:5, names_to = "ROUTE", values_to = "FLIGHTS") %>% group_by(ROUTE) %>% mutate(MOVING_MEDIAN = quantile(FLIGHTS[month(DATE) %in% month(last(DATE))-2:month(last(DATE))], probs = 0.5), NORM_FLTS = FLIGHTS/MOVING_MEDIAN)
-temp2
-
-temp2 %>% ggplot(aes(x = DATE)) +
-  geom_line(aes(y = NORM_FLTS, color = ROUTE)) +
-  theme_minimal()
+  geom_line(aes(y = (FLIGHTS), color = ROUTE))
 ```
-
-```{r}
-
-```
-
-1.2.2
 
@@
 
+This paper addressed the identification and description of disruptions, associated transformation, and recovery phenomena of the air transport network.
+The methodology will be useful to support the monitoring of the recovery levels across the globe.
+It complements the toolbox for policy makers and strategic planers to determine priority measures in support of the anticipated air travel demand recovery with the roll-out and increasing levels of vaccination.
+
 <!-- 4. Recommendations for further work. -->
 
+<!-- final words -->
+As part of this project, the requirements for an open data based flight table for the evaluation of the global air transportation network are refined. 
+The results and insights of this work will be shared with the wider effort of the ICAO GANP expert group on performance.
+
 # Acknowledgment {.unnumbered}
+
+<!-- think about providing pointer to paper repo/data vs IEEE/DASC copyright reqs
+The paper is written as a fully reproducible paper based on open datasets.
+Reproducibility and transparency will be further ensured by utilizing an open source software environment.
+In that respect, interested researchers or practitioners will be able to validate all aspects of the code, data, and published results.
+-->
 

### 2021-07-12 paper.Rmd
approach and workflow

@@
 The pressing, increasing demand for air transport registered in the last decade not only has already produced challenging delay management practices, but also fostered projections of even worse scenarios.
-EUROCONTROL (\_\_\_\_), for example, argued that delays in Europe could reach up to 20 minutes per flight in 2040, in stark contrast to the 12 minutes per flight, as registered in 2016.
+EUROCONTROL [@statfor_growth_2018], for example, argued that delays in Europe could reach up to 20 minutes per flight in 2040, in stark contrast to the 12 minutes per flight, as registered in 2016.
 
@@
 
-(ref:fig-global-covid) Reported number of cases (22. Jan 2020 through 27. May 2021) [@coronavirus_data] based on data from John Hopkins University
+(ref:fig-global-covid) Reported number of cases (22. Jan 2020 through 27. May 2021) [@krispin_coronavirus_2021] based on data from John Hopkins University [@JHU_2020]
 
@@
 ```
+
+TODO
+
+* provide some descriptive text about COVID and government reactions
+* any more key milestones relevant for this study, e.g. US/EU stop, Lufthansa/BA/KLM/Air France stopping to operate to China as of late January
 
@@
 
-A mixed-method approach, based on:
+```{r}
+# A mixed-method approach, based on:
+# 
+# a)  to answer RQ1.1, a qualitative analysis of previous models to develop acute low-demand as a disturbance
+# b)  to answer RQ1.2, a quantitative analysis of open data, to observe (or not) different levels/stages of stress/recovery, which could indicate different "more" or "less" resilience to the disturbances
+```
 
-a)  to answer RQ1.1, a qualitative analysis of previous models to develop acute low-demand as a disturbance
-b)  to answer RQ1.2, a quantitative analysis of open data, to observe (or not) different levels/stages of stress/recovery, which could indicate different "more" or "less" resilience to the disturbances
+## Research Approach and Toolset
+
+This work follows a data-driven approach and is based on the reproducibility paradigm. 
+By design open data sources/sets have been identified and used.
+The associated data analysis is conducted using the open-source software and toolkit R, RStudio, and various packages of the R-ecosystem.
+The paper and its supporting datasets or code to retrieve the data have been published and are freely available.
+
+Figure \@ref(fig:fig-research-approach) shows the overall approach workflow for this paper.
+The primary open data source for air traffic data is the global daily flight data set [@xavier_olive_2021_4893103].
+
+<<VERBATIM -- NEED TO REPHRASE THIS>>----------- 
+This  work  builds  on  the  RStudio [@rstudio]  tools  for  the  R  language [@R]  including  Git  (and  the  web-based  repository  managersGitHub  and  GitLab)  as  underlying  version  control  system.The R language was originally developed within the statisticalcommunity supporting the task of statistical reporting by pro-viding routines for the statistical computing and visualisation.Being  open  source,  the  R  community  is  actively  engagingand  sharing  related  software  packages  to  enhance  the  corefunctionality.
+Without  limiting  the  impact  of  other  packages,the  development  of  knitr  [14]  and  RMarkdown  [15],  ggplot[16]  for  visualisation,  and  the  so-called  tidyverse  packagesand  RStudio  IDE  [17]  represent  an  open  source  ecosystemfor data analysis.
+A key feature for the implementation of thereproducibility paradigm is the fact that RMarkdown supportsthe combination of text, analytical code, and visualisations ina  single  document.
+<<END VERBATIM COPY -------------------------------------->>
+
+(ref:fig-research-approach) Reproducible Research Approach
+
+```{r fig-research-approach, fig.cap="(ref:fig-research-approach)"}
+knitr::include_graphics("./figures/approach.png")
+```
 

### 2021-07-12 paper.Rmd
static impact/disturbance levels included to method section (next: sliding window by Fabio)

@@
 <!--| 1. Nature of the problem. -->
-
-This paper is heavily informed by the work of [@strohmeier_crowdsourced_2021].
 
@@
 
+The global daily flight dataset by Opensky Network identifies a flight based on a series of received messages of 15 minutes or more [@strohmeier_crowdsourced_2021].
+As the sensor network does not cover the whole globe/airspace, a consistency check is made for flights leaving the coverage area and entering again (e.g. oceanic area).
+A heuristic algorithm for the identification of destination airports is applied for descending trajectories. 
+While there are uncertainties in the data based on these heuristics, the dataset ensures detection of flights within certain geographic regions even without a positive departure and/or destination aerodrome identification.
+
 <!-- add some data set description, e.g. number of files, etc -->
+
+> TODO: develop table with months - file size - number of flights
 
@@
 as presented in Fig. \@ref(fig:resiliencebasics) (LOS: loss of service / performance, THR: associated threshold).
+
+```{r}
+# load analytic dataset
+us <- ds %>% 
+  filter(ADEP_REG == "US" | ADES_REG == "US") %>%
+  group_by(YR = year(DATE), DATE) %>% 
+  summarize(N = n(), .groups = "drop")
+
+# static thresholds, x-th percentile rounded to lower integer
+pcts   <- c(0.8, 0.6, 0.3, 0.1)
+us_ref <- us %>% filter(YR == 2019) %>%
+  pull(N) %>% quantile(probs = pcts) %>% floor()
+
+us_ref2 <- us %>% filter(YR == 2019, DATE >= lubridate::ymd("2019-07-01")) %>%
+  pull(N) %>% quantile(probs = pcts) %>% floor()
+
+max_N <- max(us$N)
+lvls  <- c("First", "Second", "Third", "Fourth", "Fifth")
+us <- us %>% 
+  mutate( THRES  = cut(N, breaks = c(max_N, us_ref, 0),  labels = lvls)
+         ,THRES2 = cut(N, breaks = c(max_N, us_ref2, 0), labels = lvls)
+         )
+```
+
+(ref:fig-us-timeline) Resilience concept applied to observed US air traffic development
+
+```{r fig-us-timeline, fig.cap="(ref:fig-us-timeline)"}
+us %>%
+  ggplot(aes(x = DATE)) +
+  geom_hline(yintercept = us_ref2, linetype = "dotdash") +
+  geom_point(aes(y = N, color = THRES2), alpha = 0.3) +
+  geom_line( aes(y = N, colour = THRES2, group = 1)) +
+  theme_minimal()
+```
+
+> TODO - further develop & include sliding window approach proposed by Fabio
 

### 2021-07-12 x0_data-preparation.Rmd
static impact/disturbance levels included to method section (next: sliding window by Fabio)

--- 2021-06-24 x0_data-preparation.Rmd
@@
 
-```{r, eval=TRUE}
+```{r, eval=TRUE, message=FALSE}
 #quick peek
@@
 ```
+
+```{r, dpi=480}
+#THRESHOLD AT ROLLING QUANTILES PER YEAR
+#Please install the "roll" package.
+#install.packages("roll")
+library(roll)
+
+ds1 <- full_dataset %>% group_by(YR = year(DATE), DATE) %>% summarize(N = n(), .groups = "drop") %>% mutate(THRESHOLD = quantile(N, probs = 0.7))
+ds1
+
+ds1 %>% ggplot(aes(x = DATE)) +
+  geom_line(aes(y = THRESHOLD)) +
+  geom_point(aes(y = N, color = N > THRESHOLD))
+```
+
+```{r}
+#CHECK WEEKLY DISTRIBUTION
+ds2 <- ds1 %>% 
+  mutate( WEEK_DAY = wday(DATE, label = TRUE)
+         ,W_END    = WEEK_DAY %in% c("Sat", "Sun"))
+uplim <- .95
+lowlim <- .20
+
+ds2 <- ds2 %>% mutate(MED = roll_quantile(N, width = 7, p = .50, min_obs = 7),
+                      UP_THR = roll_quantile(N, width = 180, p = uplim, min_obs = 7),
+                      LOW_THR = roll_quantile(N, width = 180, p = lowlim, min_obs = 7))
+ds2
+
+ds2 %>% ggplot(aes(x = DATE)) +
+  geom_col(aes(y = MED, fill = !(MED < LOW_THR | MED > UP_THR))) +
+  geom_line(aes(y = UP_THR), color = "blue") +
+  geom_line(aes(y = LOW_THR), color = "red") +
+  scale_x_date(limits = c(as.Date("2019-10-01"), as.Date("2020-12-01")), date_breaks = "30 days", date_labels = "%b\n%y") +
+  labs(y = "7-day rolling median", x = NULL, fill = "Normal operation?",
+       title = "What is normal?",
+       subtitle = "Red line - Lower threshold; Blue line - upper threshold",
+       caption = "Lower threshold set to 20th percentile of the last 180 days;\nUpper threshold set to 95th percentile of the last 180 days"
+)
+```
+
+```
+# breadcrumb ------------- need to check what this is for --------------------
+#ds2 %>% ggplot() +
+  geom_freqpoly(aes(x = N, color = W_END))
+```
+
+```{r}
+us <- full_dataset %>% 
+  filter(ADEP_REG == "US" | ADES_REG == "US") %>%
+  group_by(YR = year(DATE), DATE) %>% 
+  summarize(N = n(), .groups = "drop")
+
+# static thresholds, x-th percentile rounded to lower integer
+pcts   <- c(0.8, 0.6, 0.3, 0.1)
+us_ref <- us %>% filter(YR == 2019) %>%
+  pull(N) %>% quantile(probs = pcts) %>% floor()
+
+us_ref2 <- us %>% filter(YR == 2019, DATE >= lubridate::ymd("2019-07-01")) %>%
+  pull(N) %>% quantile(probs = pcts) %>% floor()
+
+max_N <- max(us$N)
+lvls  <- c("First", "Second", "Third", "Fourth", "Fifth")
+us <- us %>% 
+  mutate( THRES  = cut(N, breaks = c(max_N, us_ref, 0),  labels = lvls)
+         ,THRES2 = cut(N, breaks = c(max_N, us_ref2, 0), labels = lvls)
+         )
+```
+```{r}
+us %>%
+  ggplot(aes(x = DATE)) +
+ # geom_line(aes(y = THRES)) +
+  geom_point(aes(y = N, color = THRES2)) +
+  geom_line(aes(y = N, colour = THRES2, group = 1)) +
+  geom_hline(yintercept = c(max_N, us_ref2, 0))
+```

### 2021-07-12 paper.Rmd
network level assessment, combined timeline (need to work out nicer design)

@@
 library(zoo)
+library(patchwork)
 
@@
 ```{r}
-# load analytic dataset
-us <- ds %>% 
-  filter(ADEP_REG == "US" | ADES_REG == "US") %>%
+# summarise regional level
+## -------- percentile thresholds
+pcts   <- c(0.8, 0.6, 0.3, 0.1)
+
+## -------- function definition
+regional_timeline <- function(.ds, .reg, .pcts=pcts){
+us <- .ds %>% 
+  filter(ADEP_REG == .reg | ADES_REG == .reg) %>%
   group_by(YR = year(DATE), DATE) %>% 
@@
 # static thresholds, x-th percentile rounded to lower integer
-pcts   <- c(0.8, 0.6, 0.3, 0.1)
 us_ref <- us %>% filter(YR == 2019) %>%
-  pull(N) %>% quantile(probs = pcts) %>% floor()
+  pull(N) %>% quantile(probs = .pcts) %>% floor()
 
 us_ref2 <- us %>% filter(YR == 2019, DATE >= lubridate::ymd("2019-07-01")) %>%
-  pull(N) %>% quantile(probs = pcts) %>% floor()
+  pull(N) %>% quantile(probs = .pcts) %>% floor()
 
@@
          )
+return(list(us, us_ref, us_ref2))
+}
+
+## do for us
+res <- regional_timeline(ds, "US")
+us  <- res[[1]]
+us_ref2 <- res[[3]]
 ```
@@
 ```{r fig-us-timeline, fig.cap="(ref:fig-us-timeline)"}
-us %>%
+plot_timeline <- function(.df, .pct_ref){
+ p <- .df %>%
   ggplot(aes(x = DATE)) +
-  geom_hline(yintercept = us_ref2, linetype = "dotdash") +
+  geom_hline(yintercept = .pct_ref, linetype = "dotdash") +
   geom_point(aes(y = N, color = THRES2), alpha = 0.3) +
@@
   theme_minimal()
+ return(p)
+}
+
+ gg <- plot_timeline(us, us_ref2) 
+ gg
 ```
@@
 
-1.1
+## Network Level Assessment for Brazil, Europe, and the United States
+
+Based on the study dataset Fig. \@ref(fig:osndaily) provides an appreciation of the global developments.
+This paper zooms in on comparing the situation in Brazil, Europe, and the United States.
+Figure \@ref(fig:us-eu-br-timeline) shows the different network level responses to the regional/local air traffic constraints.
+
+```{r}
+us_timeline <- regional_timeline(ds, "US")
+eu_timeline <- regional_timeline(ds, "EU")
+br_timeline <- regional_timeline(ds, "BR")
+
+trim_timeline <- function(.df){
+  df <- .df %>% filter(DATE >= lubridate::ymd("2020-01-01"))
+}
+
+us_p <- plot_timeline(us_timeline[[1]] %>% trim_timeline(), us_timeline[[3]])
+eu_p <- plot_timeline(eu_timeline[[1]] %>% trim_timeline(), eu_timeline[[3]])
+br_p <- plot_timeline(br_timeline[[1]] %>% trim_timeline(), br_timeline[[3]])
+```
+
+(ref:us-eu-br-timeline) Timeline for United States, Europe, and Brazil
+
+```{r us-eu-br-timeline, fig.cap="(ref:us-eu-br-timeline)"}
+( (us_p + labs(x=NULL, y=NULL, subtitle = "United States")) + 
+  (eu_p + labs(x=NULL, y=NULL, subtitle = "Europe"       )) + 
+  (br_p + labs(x=NULL, y=NULL, subtitle = "Brazil"       ))   ) + 
+  plot_layout(guides = "collect") &
+  theme(legend.position='bottom')
+```
+
+The United States have recovered.
+
+Europe has seen an initial recovery in summer 2020 with a decline towards the holiday season 2020/2021.
+With April/May 2021 traffic increases again in Europe.
+
+Brazil also experienced the sharp decline in traffic in March 2020.
+Similar to the US, traffic in Brazil showed an initial continual increase. 
+Traffic show an alternating pattern.
+
+```
+TODO: provide some more commentary
+```
+
+## Traffic Development - Connectivity
 

### 2021-07-13 x0_data-preparation.Rmd
sort of zooming in on parameters for upper/lower limit in window size: uplim = 0.8, lowlim = 0.2, step = 60

@@
 
-```{r}
+loop deactivated as base data has been saved out and shared across the project team
+
+```{r, eval=FALSE}
 # read all files (note adpat function name)
@@
 ```
+
+```{r}
+# comparator levels
+uplim  <- .80
+lowlim <- .20
+slide  <-  60
+
+tmp <- us %>% 
+  mutate( N_sm    = roll_quantile(N, width = 7, p = .50)
+         ,MED     = roll_quantile(N, width = 14, p = .50, min_obs = 7)
+         ,UP_THR  = roll_quantile(N, width = slide, p = uplim,  min_obs = 7)
+         ,LOW_THR = roll_quantile(N, width = slide, p = lowlim, min_obs = 7)
+         ,LOW_CRIT= LOW_THR >= MED
+         ,UP_CRIT = UP_THR  <= MED
+         ,STATE   = case_when( !LOW_CRIT & !UP_CRIT ~ "LEVEL"
+                              , LOW_CRIT & !UP_CRIT ~ "CRUNCH"
+                              ,!LOW_CRIT &  UP_CRIT ~ "RECOV" 
+                              ,TRUE ~ "UNKWN")
+         ,BAND    = 4000
+         ) %>%
+  filter(DATE >= lubridate::ymd("2019-10-01"))
+
+tmp %>%
+  ggplot() +
+  geom_line(aes(x = DATE, y = N), colour = "grey80") +
+  geom_line(aes(x = DATE, y = N_sm, colour = THRES2, group = "RQ"), size = 2) +
+  geom_line(aes(x = DATE, y = UP_THR), color = "blue") +
+  geom_line(aes(x = DATE, y = LOW_THR), color = "red") +
+  geom_line(aes(x = DATE, y = MED), color = "green") +
+  geom_line(aes(x = DATE, y = BAND, group= "1"), colour = if_else(tmp$LOW_CRIT, "red","green"), size = 4) +
+  geom_line(aes(x = DATE, y = BAND-500, group= "1"), colour = if_else(tmp$UP_CRIT, "red","green"), size = 4) +
+  geom_line(aes(x = DATE, y = BAND-1500, group= "1", colour = STATE), size = 4) +
+  theme_minimal() +
+  theme(legend.position = "none")
+```

### 2021-07-13 x0_data-preparation.Rmd
added new_scale_colour

@@
 knitr::opts_chunk$set(echo = TRUE)
+#install.packages("ggnewscale")
 library(tidyverse)
@@
 library(zoo)
+library(ggnewscale)
 
@@
 
+###---------Rainer, the "new_scale_colour()" kinda resets the scale, and you can use another one for the next geom_lines.
+In the example below, I left the legend.position = "bottom" (instead of "NULL"), just to see the colors------###
 tmp %>%
@@
   geom_line(aes(x = DATE, y = MED), color = "green") +
-  geom_line(aes(x = DATE, y = BAND, group= "1"), colour = if_else(tmp$LOW_CRIT, "red","green"), size = 4) +
+  scale_color_brewer(type = "seq") +
+  new_scale_colour() +
+  geom_line(aes(x = DATE, y = BAND, group= "1"), colour = if_else(tmp$LOW_CRIT, "red","green"), size = 10) +
   geom_line(aes(x = DATE, y = BAND-500, group= "1"), colour = if_else(tmp$UP_CRIT, "red","green"), size = 4) +
   geom_line(aes(x = DATE, y = BAND-1500, group= "1", colour = STATE), size = 4) +
+  scale_colour_brewer(type = "qual") +
   theme_minimal() +
-  theme(legend.position = "none")
+  theme(legend.position = "bottom")
 ```

### 2021-07-13 x0_data-preparation.Rmd
my tweaks for the airport level

--- 2021-07-13 x0_data-preparation.Rmd
@@
   geom_line(aes(x = DATE, y = BAND-500, group= "1"), colour = if_else(tmp$UP_CRIT, "red","green"), size = 4) +
-  geom_line(aes(x = DATE, y = BAND-1500, group= "1", colour = STATE), size = 4) +
+  geom_line(aes(x = DATE, y = BAND-1500, group= "2", colour = STATE), size = 4) +
   theme_minimal() +
@@
 ```
+
+# Airport Level Considerations
+
+top-5 in the whole sample, confirms preference :)
+
+```{r}
+full_dataset %>% 
+  group_by(ADEP_REG, ADEP) %>% 
+  summarise(N = n()) %>% 
+  slice_max(order_by = N, n = 5)
+```
+
+```{r}
+sbgr <- full_dataset %>% filter(ADEP == "SBGR")
+```
+
+```{r}
+apt_timeline <- function(.df){
+  df <- .df %>% group_by(DATE) %>%
+    summarise(N_DEP = n()
+              ,DEP_DOM = sum(ADEP_CTRY == ADES_CTRY)
+              ,DEP_REG = sum(ADEP_REG == ADES_REG))
+}
+
+plot_timeline <- function(.df){
+  df <- .df %>% 
+  mutate(N_DEP_SM = roll_quantile(N_DEP, width = 7, p = .50))
+
+  p <- df %>%  
+  ggplot() +
+  geom_line(aes(x = DATE, y = N_DEP), colour = "grey60") +
+  geom_line(aes(x = DATE, y = N_DEP_SM)) +
+  scale_x_date( limits = c(as.Date("2019-10-01"), as.Date("2021-06-01"))
+               ,date_breaks = "60 days", date_labels = "%b\n%y") +
+  theme_minimal()
+
+  return(p)
+}
+
+sbgr %>% apt_timeline() %>% plot_timeline()
+```
+
+```{r}
+sbgr_tl %>%
+  ggplot() +
+  geom_point(aes(x = DATE, y = DEP_DOM / N_DEP)) +
+  theme_minimal()
+```
+```{r}
+dep_vs_domshare <- function(.df){
+  p <- .df %>%
+  ggplot() +
+  geom_point(aes(x = N_DEP, y = DEP_REG / N_DEP, colour = as.factor(year(DATE)))) +
+  labs(colour = NULL) + 
+  scale_y_continuous(labels = scales::percent_format(accuracy = 1)
+                     ,expand = c(0,0)) +
+  theme_minimal() +
+  theme(legend.position = c(0.8, 0.8))
+return(p)
+}
+
+sbgr_tl %>% dep_vs_domshare()
+```
+
+```{r}
+check_apt <- function(.mvts, .apt){
+  apt_mvts <- .mvts %>%
+    filter(ADEP == .apt) %>% 
+    apt_timeline()
+
+  p1 <- apt_mvts  %>% plot_timeline()
+  p2 <- apt_mvts %>% dep_vs_domshare()
+
+  p <- (p1 + labs(subtitle = .apt)) + p2
+return(p)
+}
+
+full_dataset %>% check_apt("EGLL")
+```
+
+```{r}
+full_dataset %>% check_apt("EDDF")
+```
+
+```{r}
+full_dataset %>% check_apt("LFPG")
+```
+
+```{r}
+full_dataset %>% check_apt("KATL")
+```
+
+```{r}
+full_dataset %>% check_apt("KORD")
+```
+
+```{r}
+full_dataset %>% check_apt("KJFK")
+```
+
+```{r}
+full_dataset %>% check_apt("EHAM")
+```
+
+```{r}
+full_dataset %>% check_apt("EHAM")
+```

### 2021-07-13 x0_data-preparation.Rmd

@@
 library(patchwork)
+library(ggnewscale)
 
@@
   geom_line(aes(x = DATE, y = MED), color = "green") +
-  geom_line(aes(x = DATE, y = BAND, group= "1"), colour = if_else(tmp$LOW_CRIT, "red","green"), size = 4) +
+  scale_color_brewer(type = "seq") +
+  new_scale_colour() +
+  geom_line(aes(x = DATE, y = BAND, group= "1"), colour = if_else(tmp$LOW_CRIT, "red","green"), size = 10) +
   geom_line(aes(x = DATE, y = BAND-500, group= "1"), colour = if_else(tmp$UP_CRIT, "red","green"), size = 4) +
-  geom_line(aes(x = DATE, y = BAND-1500, group= "2", colour = STATE), size = 4) +
+  geom_line(aes(x = DATE, y = BAND-1500, group= "1", colour = STATE), size = 4) +
+  scale_colour_brewer(type = "qual") +
   theme_minimal() +
-  theme(legend.position = "none")
+  theme(legend.position = "bottom")
 ```

### 2021-07-13 x0_data-preparation.Rmd
got weird push error - inserted a line to force another commit/push

@@
 #install.packages("ggnewscale")
+
 library(tidyverse)

### 2021-07-14 x0_data-preparation.Rmd
just a small correction to read the next lines

@@
 ```{r}
+sbgr_tl <- sbgr %>% apt_timeline()
+
 sbgr_tl %>%

### 2021-07-16 paper.Rmd
a few corrections and added some government reactions

@@
 
-For many years, many concerns of the global air traffic management community has been directed to the evident problem of imbalances between capacity and demand.
+For several years, many concerns of the global air traffic management community has been directed to the evident problem of imbalances between capacity and demand.
 The pressing, increasing demand for air transport registered in the last decade not only has already produced challenging delay management practices, but also fostered projections of even worse scenarios.
@@
 As a result, air navigation providers collected less fees for their services, due to significantly fewer flights.
-In the operational perspective, pilots and air traffic controllers practiced less. The problems and obstacles developed into many other dimensions.
+In the operational perspective, pilots and air traffic controllers practiced less. The problems and obstacles cascaded into many other dimensions.
 
@@
 A variety of studies showed the interplay or consequences of the travel constraints. 
-To date, lower attention was given to the inherent change in terms of air transport services
+To date, lower attention was given to the inherent change in terms of air transport services.
 Hence, the current scenario is a proper moment to further investigate the concept of resilience.
@@
 
-* provide some descriptive text about COVID and government reactions
+* provide some descriptive text about COVID and government reactions (DONE BELOW!):
+
+For example, in March 12, 2020, the United States established the suspension of most travels from the European Schengen area (BBC, 2020).
+A few days later, European Comission President Ursula von der Leyen announced a similar measure, proposing for EU Member States to apply restrictions on non-essential travel.
+By that time, several countries, such as Austria, Germany, and Poland had already implemented border check controls, while Portugal, Italy, and Spain have imposed lockdowns (Nielsen, 2020).
+At the same time, Australia imposed a severe arrival block on all non-citizens and non-residents at its airports (Worthington & Snape, 2020).
+In South America, Brazil, Chile, Argentina and other countries also followed the trend and closed their borders, totally or partially (France24, 2020).
+As a result, an unparalleled air traffic drop was seen worldwide.
+
+BBC. (2020, March 12). Coronavirus: Trump suspends travel from Europe to US. https://www.bbc.com/news/world-us-canada-51846923
+
+France24. (2020, March 16). Chile closes borders as Latam grounds flights over pandemic. https://www.france24.com/en/20200316-chile-closes-borders-as-latam-grounds-flights-over-pandemic
+
+Nielsen, N. (2020, March 17). Commission proposes 'non-essential travel' EU entry ban. EU Observer. https://euobserver.com/coronavirus/147764
+
+Worthington, B., & Snape, J.
+(2020, March 19).
+Australia blocks arrival of all non-citizens, non-residents in expanded coronavirus travel ban.
+ABC News.
+https://www.abc.net.au/news/2020-03-19/coronavirus-non-resident-travel-ban-australia/12071640
+
 * any more key milestones relevant for this study, e.g. US/EU stop, Lufthansa/BA/KLM/Air France stopping to operate to China as of late January

### 2021-07-16 paper.Rmd
added a figure and the references for the paragraph

@@
 In South America, Brazil, Chile, Argentina and other countries also followed the trend and closed their borders, totally or partially (France24, 2020).
-As a result, an unparalleled air traffic drop was seen worldwide.
+As a result, an unparalleled air traffic decline was seen worldwide (ICAO, 2021).
 
-BBC. (2020, March 12). Coronavirus: Trump suspends travel from Europe to US. https://www.bbc.com/news/world-us-canada-51846923
+```{r, out.width='50%', fig.align='center'}
+### It's not really open data, but shows the unprecedented decline mentioned above.###
+#Use this only if you want...#
 
-France24. (2020, March 16). Chile closes borders as Latam grounds flights over pandemic. https://www.france24.com/en/20200316-chile-closes-borders-as-latam-grounds-flights-over-pandemic
+knitr::include_graphics('./figures/unprecedented_decline.jpg')
 
-Nielsen, N. (2020, March 17). Commission proposes 'non-essential travel' EU entry ban. EU Observer. https://euobserver.com/coronavirus/147764
+#Please add the references below to the bib:
 
-Worthington, B., & Snape, J.
+#BBC. (2020, March 12). Coronavirus: Trump suspends travel from Europe to US. https://www.bbc.com/news/world-us-canada-51846923
+
+#France24. (2020, March 16). Chile closes borders as Latam grounds flights over pandemic. https://www.france24.com/en/20200316-chile-closes-borders-as-latam-grounds-flights-over-pandemic
+
+#International Civil Aviation Organization.
+(2021, July 13).
+Effects of novel Coronavirus (COVID‐19) on civil aviation:Economic impact analysis.
+https://www.icao.int/sustainability/Documents/COVID-19/ICAO_Coronavirus_Econ_Impact.pdf
+
+#Nielsen, N. (2020, March 17). Commission proposes 'non-essential travel' EU entry ban. EU Observer. https://euobserver.com/coronavirus/147764
+
+#Worthington, B., & Snape, J.
 (2020, March 19).
@@
 https://www.abc.net.au/news/2020-03-19/coronavirus-non-resident-travel-ban-australia/12071640
+
+```
 

### 2021-07-16 paper.Rmd
added two paragraphs ref political and operational interests on resilience.

@@
 Recently, the term is also used frequently on the political and strategic level.
-Policymakers, operational experts, and academes concur that the concept of resilience plays a major role when addressing and assessing the extent to which organisations and systems are prepared and capable to respond to and recover from disruptions.
+Policymakers, operational experts, and academics concur that the concept of resilience plays a major role when addressing and assessing the extent to which organisations and systems are prepared and capable to respond to and recover from disruptions.
 
@@
 
--   political level: ADD SOME DESCRIPTION
+-   political level: The continuity and sustainability of a stable air traffic system supports and affects other dimensions of political interest, such as commercial relations, tourism industry, public health policies, among others.
+As a result, at a political level, it is of the utmost importance that the air navigation system remains functional.
+In that understanding, political acts were necessary in face of the pandemic crisis effects on aviation.
+For example, many governments injected financial support in the airline sector in order to help  them survive the worst periods.
+In addition, measures such as tax deferrals to air navigation services providers were also adopted.
+Therefore, understanding how the air navigation system can exist in a more resilient manner is certainly beneficial for many political agents.
 
--   operational level: ADD SOME DESCRIPTION
+-   operational level: While efficiency, delay management, and capacity constraints were previously the main problems affecting resilience, now the concept must be broadened to an inverted scenario.
+Airlines, ANSPs, airports, and other stakeholders in the industry must remain and manage resilience not only in order to recover from disruptions caused by excessive demand, but also from insufficient demand.
+If delays are no longer the main threat, others arise, such as insufficient funding, lack of training, and traffic unpredictability.
 

### 2021-07-16 paper.Rmd
transformed the bullets about resilience in readable paragraphs

@@
 Within air navigation, the term 'resilience' has been picked up by several communities, both operational and scientific.
+The concept first appeared in this context as a definition proposed by EUROCONTROL (2009): "Resilience is the intrinsic ability of a system to adjust its functioning prior to, during, or following changes and disturbances, so that it can sustain required operations under both expected and unexpected conditions".
+Later, Gluchshenko (2012) proposed a widened view, including definitions for resilience, robustness, disturbance, stress, and perturbation.
+In addition, the work marks the first proposition for a framework of different levels of stress and perturbations, as well as a proposition of potential metrics for resilience (both quantitative and qualitative).
+On a later work, Gluchshenko (2013) repeats the previous ideas and adds a performance-based approach as well as an algorithm to investigate resilience.
 
-EUROCONTROL (2009): first definition of resilience in ATM context -- "Resilience is the intrinsic ability of a system to adjust its functioning prior to, during, or following changes and disturbances, so that it can sustain required operations under both expected and unexpected conditions".
-
-Gluchshenko (2012):
-
-Definitions for Resilience, robustness, disturbance, stress, and perturbation Proposition for a framework of different levels of stress/perturbations Proposition of metrics for resilience (both quantitative and qualitative)
-
-Gluchshenko (2013): repeats the previous ideas and adds a performance-based approach as well as an algorithm to investigate resilience
-
-Project Resilience 2050 (Jun/2012 + 43 months) -- includes the previous definitions and other technical tasks.
-However, it evolves the way to measure resilience.
-Now, not only the time of deviation and time of recovery is considered.
-The project measures it as the relative difference of rate of delays correlation, or R = (ax1 -- dx1)/dx1 -- it has no unit, it's the difference between two pearson correlations.
-
-Koelle (2015): proposes to address resilience as a situation management and state-oriented problem.
-Through two case studies, argued that "there is a lack of fit of the current operational ANS performance indicators to address impact of disruptions as they are primarily based on actual timestamps or transition times."
+Those were the basic concepts that supported further works regarding resilience within the air navigation context.
+For example, the Project Resilience 2050 addresed the previous definitions and other technical tasks.
+The project ran from June 2012 and lasted 43 months, evolving the way to measure resilience.
+For the project, not only the time of deviation and time of recovery is considered.
+Instead, it measures resilience as the relative difference of rate of delays correlation, or R = (ax1 -- dx1)/dx1.
+As a difference between two pearson correlations, it has no unit adopted.
+Finally, Koelle (2015) proposed to address resilience as a situation management and state-oriented problem.
+Through two case studies, the author argued that "there is a lack of fit of the current operational ANS performance indicators to address impact of disruptions as they are primarily based on actual timestamps or transition times."
 

### 2021-07-16 x0_data-preparation.Rmd
total control of colors in moving quantile graphs

@@
 In the example below, I left the legend.position = "bottom" (instead of "NULL"), just to see the colors------###
-tmp %>%
+Glass <- rgb(0, 0, 0, max = 255, alpha = 0, names = "transparent")
+
+ResilienceMovingBands <- tmp %>%
   ggplot() +
@@
   geom_line(aes(x = DATE, y = LOW_THR), color = "red") +
-  geom_line(aes(x = DATE, y = MED), color = "green") +
-  scale_color_brewer(type = "seq") +
+  #geom_line(aes(x = DATE, y = MED), color = "forestgreen", size = 2) +
+  #scale_color_brewer(type = "qual") +
   new_scale_colour() +
-  geom_line(aes(x = DATE, y = BAND, group= "1"), colour = if_else(tmp$LOW_CRIT, "red","green"), size = 10) +
-  geom_line(aes(x = DATE, y = BAND-500, group= "1"), colour = if_else(tmp$UP_CRIT, "red","green"), size = 4) +
-  geom_line(aes(x = DATE, y = BAND-1500, group= "1", colour = STATE), size = 4) +
-  scale_colour_brewer(type = "qual") +
+  geom_line(aes(x = DATE, y = BAND-100, group= "1"), colour = if_else(tmp$LOW_CRIT, "red",Glass), size = 4) +
+  geom_line(aes(x = DATE, y = BAND+16000, group= "1"), colour = if_else(tmp$UP_CRIT, "blue", Glass), size = 4) +
+  geom_line(aes(x = DATE, y = BAND-4000, group= "1", colour = STATE), size = 10) +
+  scale_colour_manual(values = c("darkred", Glass, "darkblue")) +
   theme_minimal() +
   theme(legend.position = "bottom")
+ResilienceMovingBands
 ```

### 2021-07-21 x0_data-preparation.Rmd
geo coding

--- 2021-07-13 x0_data-preparation.Rmd
@@
 knitr::opts_chunk$set(echo = TRUE)
-#install.packages("ggnewscale")
-
 library(tidyverse)
@@
 library(zoo)
-library(patchwork)
-library(ggnewscale)
+library(sf)
+library(countrycode)
 
@@
             ,ADES = destination
+            ,ICAO24 = icao24
             ,TYPE = as.factor(typecode)
@@
             #,ACFT_ID = aircraft_uid    # throws an error in today's data download
+            ,LAT1 = latitude_1
+            ,LON1 = longitude_1
+            ,ALT1 = altitude_1
+            ,LAT2 = latitude_2
+            ,LON2 = longitude_2
+            ,ALT2 = altitude_2
             )
@@
 
+# ====================== geo coding stuff ===================================
+
+#' utility function to geocode lat lon positions - read shape files for geocoding
+#' 
+load_shapefile_for_geo_ref <- function(){
+  EEZ_land_shp <- list.files("../__DATA/xWorld_EEZ_and_land_v3", pattern = ".shp$", full.names = TRUE)
+  eez_land <- sf::read_sf(EEZ_land_shp) %>% select(ISO_TER1)
+}
+
+#' utility function to coerce
+
+# ====================== wrap all prep in one function ===================
+
 prep_base_data <- function(.fn = file_names, ...){
@@
 ds <- ds %>% append_ad_ctry()
+
+ds_ok  <- ds %>% filter(!is.na(ADEP), !is.na(ADES))
+ds_nok <- ds %>% filter( is.na(ADEP) | is.na(ADES))
+
 ds <- ds %>% append_region()
@@
 
-ds <- ds %>% drop_na()  # drop all ~ assumed to be clean
+#ds <- ds %>% drop_na()  # drop all ~ assumed to be clean
 
@@
 
-loop deactivated as base data has been saved out and shared across the project team
-
-```{r, eval=FALSE}
+```{r}
 # read all files (note adpat function name)
@@
 
-```{r, eval=TRUE, message=FALSE}
+```{r, eval=TRUE}
 #quick peek
@@
   geom_point(aes(y = N, color = N > THRESHOLD))
-```
 
-```{r}
 #CHECK WEEKLY DISTRIBUTION
-ds2 <- ds1 %>% 
-  mutate( WEEK_DAY = wday(DATE, label = TRUE)
-         ,W_END    = WEEK_DAY %in% c("Sat", "Sun"))
+ds2 <- ds1 %>% mutate(WEEK_DAY = wday(DATE, label = TRUE), W_END = WEEK_DAY %in% c("Sat", "Sun"))
 uplim <- .95
@@
 )
-```
 
-```
-# breadcrumb ------------- need to check what this is for --------------------
 #ds2 %>% ggplot() +
@@
 ```
-
-```{r}
-us <- full_dataset %>% 
-  filter(ADEP_REG == "US" | ADES_REG == "US") %>%
-  group_by(YR = year(DATE), DATE) %>% 
-  summarize(N = n(), .groups = "drop")
-
-# static thresholds, x-th percentile rounded to lower integer
-pcts   <- c(0.8, 0.6, 0.3, 0.1)
-us_ref <- us %>% filter(YR == 2019) %>%
-  pull(N) %>% quantile(probs = pcts) %>% floor()
-
-us_ref2 <- us %>% filter(YR == 2019, DATE >= lubridate::ymd("2019-07-01")) %>%
-  pull(N) %>% quantile(probs = pcts) %>% floor()
-
-max_N <- max(us$N)
-lvls  <- c("First", "Second", "Third", "Fourth", "Fifth")
-us <- us %>% 
-  mutate( THRES  = cut(N, breaks = c(max_N, us_ref, 0),  labels = lvls)
-         ,THRES2 = cut(N, breaks = c(max_N, us_ref2, 0), labels = lvls)
-         )
-```
-```{r}
-us %>%
-  ggplot(aes(x = DATE)) +
- # geom_line(aes(y = THRES)) +
-  geom_point(aes(y = N, color = THRES2)) +
-  geom_line(aes(y = N, colour = THRES2, group = 1)) +
-  geom_hline(yintercept = c(max_N, us_ref2, 0))
-```
-
-```{r}
-# comparator levels
-uplim  <- .80
-lowlim <- .20
-slide  <-  60
-
-tmp <- us %>% 
-  mutate( N_sm    = roll_quantile(N, width = 7, p = .50)
-         ,MED     = roll_quantile(N, width = 14, p = .50, min_obs = 7)
-         ,UP_THR  = roll_quantile(N, width = slide, p = uplim,  min_obs = 7)
-         ,LOW_THR = roll_quantile(N, width = slide, p = lowlim, min_obs = 7)
-         ,LOW_CRIT= LOW_THR >= MED
-         ,UP_CRIT = UP_THR  <= MED
-         ,STATE   = case_when( !LOW_CRIT & !UP_CRIT ~ "LEVEL"
-                              , LOW_CRIT & !UP_CRIT ~ "CRUNCH"
-                              ,!LOW_CRIT &  UP_CRIT ~ "RECOV" 
-                              ,TRUE ~ "UNKWN")
-         ,BAND    = 4000
-         ) %>%
-  filter(DATE >= lubridate::ymd("2019-10-01"))
-
-###---------Rainer, the "new_scale_colour()" kinda resets the scale, and you can use another one for the next geom_lines.
-In the example below, I left the legend.position = "bottom" (instead of "NULL"), just to see the colors------###
-tmp %>%
-  ggplot() +
-  geom_line(aes(x = DATE, y = N), colour = "grey80") +
-  geom_line(aes(x = DATE, y = N_sm, colour = THRES2, group = "RQ"), size = 2) +
-  geom_line(aes(x = DATE, y = UP_THR), color = "blue") +
-  geom_line(aes(x = DATE, y = LOW_THR), color = "red") +
-  geom_line(aes(x = DATE, y = MED), color = "green") +
-  scale_color_brewer(type = "seq") +
-  new_scale_colour() +
-  geom_line(aes(x = DATE, y = BAND, group= "1"), colour = if_else(tmp$LOW_CRIT, "red","green"), size = 10) +
-  geom_line(aes(x = DATE, y = BAND-500, group= "1"), colour = if_else(tmp$UP_CRIT, "red","green"), size = 4) +
-  geom_line(aes(x = DATE, y = BAND-1500, group= "1", colour = STATE), size = 4) +
-  scale_colour_brewer(type = "qual") +
-  theme_minimal() +
-  theme(legend.position = "bottom")
-```
-
-# Airport Level Considerations
-
-top-5 in the whole sample, confirms preference :)
-
-```{r}
-full_dataset %>% 
-  group_by(ADEP_REG, ADEP) %>% 
-  summarise(N = n()) %>% 
-  slice_max(order_by = N, n = 5)
-```
-
-```{r}
-sbgr <- full_dataset %>% filter(ADEP == "SBGR")
-```
-
-```{r}
-apt_timeline <- function(.df){
-  df <- .df %>% group_by(DATE) %>%
-    summarise(N_DEP = n()
-              ,DEP_DOM = sum(ADEP_CTRY == ADES_CTRY)
-              ,DEP_REG = sum(ADEP_REG == ADES_REG))
-}
-
-plot_timeline <- function(.df){
-  df <- .df %>% 
-  mutate(N_DEP_SM = roll_quantile(N_DEP, width = 7, p = .50))
-
-  p <- df %>%  
-  ggplot() +
-  geom_line(aes(x = DATE, y = N_DEP), colour = "grey60") +
-  geom_line(aes(x = DATE, y = N_DEP_SM)) +
-  scale_x_date( limits = c(as.Date("2019-10-01"), as.Date("2021-06-01"))
-               ,date_breaks = "60 days", date_labels = "%b\n%y") +
-  theme_minimal()
-
-  return(p)
-}
-
-sbgr %>% apt_timeline() %>% plot_timeline()
-```
-
-```{r}
-sbgr_tl %>%
-  ggplot() +
-  geom_point(aes(x = DATE, y = DEP_DOM / N_DEP)) +
-  theme_minimal()
-```
-```{r}
-dep_vs_domshare <- function(.df){
-  p <- .df %>%
-  ggplot() +
-  geom_point(aes(x = N_DEP, y = DEP_REG / N_DEP, colour = as.factor(year(DATE)))) +
-  labs(colour = NULL) + 
-  scale_y_continuous(labels = scales::percent_format(accuracy = 1)
-                     ,expand = c(0,0)) +
-  theme_minimal() +
-  theme(legend.position = c(0.8, 0.8))
-return(p)
-}
-
-sbgr_tl %>% dep_vs_domshare()
-```
-
-```{r}
-check_apt <- function(.mvts, .apt){
-  apt_mvts <- .mvts %>%
-    filter(ADEP == .apt) %>% 
-    apt_timeline()
-
-  p1 <- apt_mvts  %>% plot_timeline()
-  p2 <- apt_mvts %>% dep_vs_domshare()
-
-  p <- (p1 + labs(subtitle = .apt)) + p2
-return(p)
-}
-
-full_dataset %>% check_apt("EGLL")
-```
-
-```{r}
-full_dataset %>% check_apt("EDDF")
-```
-
-```{r}
-full_dataset %>% check_apt("LFPG")
-```
-
-```{r}
-full_dataset %>% check_apt("KATL")
-```
-
-```{r}
-full_dataset %>% check_apt("KORD")
-```
-
-```{r}
-full_dataset %>% check_apt("KJFK")
-```
-
-```{r}
-full_dataset %>% check_apt("EHAM")
-```
-
-```{r}
-full_dataset %>% check_apt("EHAM")
-```

### 2021-07-21 paper.Rmd

@@
 The concept of resilience (and robustness) is intimately linked with the construct of disruptions.
+Any given disturbance forces a disruption in the level of service.
+Following Gluchshenko's work (2012), the longer the system takes to recover to a previously unperturbed state, the less resilient the system is.
+In complement, from Koeller's work (2015), the deeper the level of disruption, also less resilient the system is.
+As a result, a possible interpretation for resilience metrics is found considering both the duration of the disrupted service performance, and its disruption level - or how much a system level of service is affected.
 
@@
 
-Conceptually, resilience R can be measured as the observed loss in quality of service (performance) over the time to recovery, t1-t2, for a certain level of disruption.
+Consequently, resilience R can be conceptually measured as the observed loss in quality of service (performance) over the time to recovery, t1-t2, for a certain level of disruption.
 Thus, mathematically, this represents the area covered by
@@
 > TODO - further develop & include sliding window approach proposed by Fabio
+
+However, an evident question arises from this approach: what is the proper threshold for the service to be considered disrupted?
+Certain that, for some indicators, this could be a fixed value, this paper proposes an alternative method.
+The disruption threshold can be viewed as an adptive function of the recent traffic levels.
+As a result, the disruption threshold level would not be a fixed value, but a variable one, taking into account the recent trends to define a normal band based on moving windows.
+The example below shows this potential approach.
+```{r}
+#-------Remember to load and run "x0_data_preparation.Rmd". The object below is produced there!--------#
+ResilienceMovingBands +
+  theme(legend.position = "right")
+```
+
+Instead of a horizontal line (fixed value), the disruption threshold adapts according to the recent trends in traffic.
+In the figure above, the red line is the disruption threshold, defined as  the `r 100*lowlim`th percentile of the daily traffic, considered the previous `r slide` days.
+Naturally, those are empirically suggested values that could be adapted according to any particular needs.
+In addition, the blue line is set to be a potential indicator of disruptions in the opposite direction, i.e.
+a positive disruption, caused by an abnormal excessive demand.
 

### 2021-07-21 x0_data-preparation.Rmd

--- 2021-07-16 x0_data-preparation.Rmd
@@
 library(zoo)
+library(sf)
+library(countrycode)
 library(patchwork)

### 2021-07-21 x1_covid.Rmd

---
title: "COVID"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

covidregionaldata := one of the many packages dealing with pre-packaging COVID data.
Potential sources WHO (often cited as unreliable), ECDC (European Center of Disease Control - changed recently to weekly data - provides R tailored download), and JHU (kicked off data collection/harmonisation effort, often cited/used)

```{r, message=FALSE}
library(covidregionaldata)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
```

```{r}
covid_jhu <- get_national_data(source = "jhu")
covid_who <- get_national_data(source = "who")
covid_ecdc<- get_national_data(source = "ecdc")
```

```{r}
global_jhu <- covid_jhu %>%
  group_by(date) %>%
  summarise( cases_total     = sum(cases_total, na.rm = TRUE)
            ,deaths_total    = sum(deaths_total, na.rm = TRUE)
            ,recovered_total = sum(recovered_total, na.rm = TRUE))
```

```{r}
global_jhu %>% 
  pivot_longer(cols = 2:4, names_to = "category", values_to = "counts") %>%

ggplot(aes(x = date, y = counts, fill = category)) +
  geom_area() +
  labs(title = "JHU ???")
```

```{r}
global_who <- covid_who %>%
  group_by(date) %>%
  summarise( cases_total     = sum(cases_total, na.rm = TRUE)
            ,deaths_total    = sum(deaths_total, na.rm = TRUE)
            ,recovered_total = sum(recovered_total, na.rm = TRUE))
```

```{r}
global_who %>% 
  pivot_longer(cols = 2:4, names_to = "category", values_to = "counts") %>%

ggplot(aes(x = date, y = counts, fill = category)) +
  geom_area() +
  labs(title = "WHO - no recovered cases")
```

```{r}
global_ecdc <- covid_ecdc %>%
  group_by(date) %>%
  summarise( cases_total     = sum(cases_total, na.rm = TRUE)
            ,deaths_total    = sum(deaths_total, na.rm = TRUE)
            ,recovered_total = sum(recovered_total, na.rm = TRUE))
```

```{r}
global_ecdc %>% 
  pivot_longer(cols = 2:4, names_to = "category", values_to = "counts") %>%

ggplot(aes(x = date, y = counts, fill = category)) +
  geom_area() +
  labs(title = "ECDC")
```

test another COVID19 project

```{r}
#install.packages("coronavirus")
# library(coronavirus)
# ---------- it is better to pull in latest changes by calling without library 
coronavirus::update_dataset()  # pulls latest (data)version
library(coronavirus)           # then call library

covid <- coronavirus %>% as_tibble()
```

```{r}
global_cov <- covid %>% 
  group_by(type, date) %>%
  summarise(total_cases = sum(cases), .groups = "drop") %>%
  pivot_wider(names_from = type, values_from = total_cases) %>%
  arrange(date) %>%
  mutate(active = confirmed - death - recovered) %>%
  mutate(active_total = cumsum(active),
                recovered_total = cumsum(recovered),
                death_total = cumsum(death))

global_cov
```

```{r}
tmp <- global_cov %>%
  select(date, contains("total")) %>%
  pivot_longer(cols = 2:4, names_to = "category", values_to = "counts")

tmp %>%

ggplot(aes(x = date, y = counts, fill = category)) +
  geom_area() +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(legend.position = c(0.2, 0.3)) +
  scale_fill_manual( name = "cases"
                    ,values = c("orange","red", "green")
                    ,labels = c("active", "death","recovered")) +
  scale_y_continuous(labels = scales::label_number_si())
```

```{r}
conf_df <- covid %>% 
  filter(type == "confirmed") %>%
  group_by(country) %>%
  summarise(total_cases = sum(cases)) %>%
  arrange(-total_cases) %>%
  mutate(
     share  = round( 100 * (total_cases / sum(total_cases)))
    ,parents = "Confirmed cases") %>%
  ungroup() 
```

```{r}
# plotly based treemap
library(plotly)
 plot_ly(data = conf_df,
          type= "treemap",
          values = ~total_cases,
          labels= ~ country,
          parents=  ~parents,
          domain = list(column=0),
          name = "Confirmed",
          textinfo="label+value+percent parent")
```

```{r}
# install.packages("treemapify")
# for options and how to use check out: https://r-charts.com/part-whole/treemapify/

library(treemapify)

ggplot( data = conf_df
       , aes( area  = total_cases, fill = as.factor(share), colour = I("white")
             ,label = paste0(country, "\n", total_cases, "\n", share, "%")
             )
       ) +
  geom_treemap(start = "topleft") +
  geom_treemap_text(colour = "white", start = "topleft") +
  theme(legend.position = "none") +
  #scale_fill_brewer(palette = "Blues")
  scale_fill_viridis_d()
```

### 2021-07-21 x0_data-preparation.Rmd
geocoding for data cleaning and ADEP-ADES assignment

@@
 #' utility function to coerce
+pts_latlon_sf <- function(.df ,.lat, .lon, .drop=FALSE){
+  check <- nrow(.df)
+  df <- .df %>%
+    filter(!is.na(.data[[.lat]]))%>%    # index to character vector ?!!
+    sf::st_as_sf(coords = c(.lon, .lat), crs=4326, remove = .drop)
+
+  if(check > nrow(df)) warning(paste0("\nremoved ", check - nrow(df), " flights with NA in lat/lon."))
+
+  return(df)
+}
+
+#' identify territory based on lat/lon
+geo_sfpts <- function(.sfpts, .shape){
+  pts <- .sfpts %>%
+    sf::st_join(.shape)
+  return(pts)
+}
 
@@
 ds <- ds %>% append_ad_ctry() 
+ds <- ds %>% append_region()
 
-ds_ok  <- ds %>% filter(!is.na(ADEP), !is.na(ADES))
+ds_ok  <- ds %>% filter(!is.na(ADEP), !is.na(ADES)) %>% select(ADEP:CALL)
 ds_nok <- ds %>% filter( is.na(ADEP) | is.na(ADES))
 
-ds <- ds %>% append_region()
+ds_nok_adep <- ds_nok %>% filter(is.na(ADEP)) %>% 
+    pts_latlon_sf("LAT1","LON1") %>% 
+    geo_sfpts(eez_land) %>% 
+    mutate(ISO2 = countrycode(ISO_TER1, origin = "iso3c", destination = "iso2c")
+           ,ADEP     = case_when(ALT1 <= 5000 ~ 'XLOX', ALT1 > 5000 ~ 'XHIX', TRUE ~ as.character(NA))
+           ,ADEP_CTRY= case_when( ADEP == "XHIX" ~ "Other"
+                                 ,ADEP == "XLOX" & !is.na(ISO2) ~ ISO2
+                                 ,ADEP == "XLOX" & is.na(ISO2) ~ ADES_CTRY
+                                 ,TRUE ~ as.character(NA))
+           )
+
+ok <- ds_nok_adep %>% filter(!is.na(ADES))  # check there are a few apts with no country
+nok<- ds_nok_adep %>% filter( is.na(ADES)) %>%
+  sf::st_drop_geometry() %>% rename(ISO3_ADEP = ISO_TER1, ISO2_ADEP = ISO2) %>%
+  pts_latlon_sf("LAT2","LON2") %>% 
+  geo_sfpts(eez_land) %>% 
+    mutate(ISO2 = countrycode(ISO_TER1, origin = "iso3c", destination = "iso2c")
+           ,ADES     = case_when(ALT2 <= 5000 ~ 'XLOX', ALT2 > 5000 ~ 'XHIX', TRUE ~ as.character(NA))
+           ,ADES_CTRY= case_when( ADES == "XHIX" ~ "Other"
+                                 ,ADES == "XLOX" & !is.na(ISO2) ~ ISO2
+                                 ,ADES == "XLOX" & is.na(ISO2) ~ ADEP_CTRY
+                                 ,TRUE ~ as.character(NA))
+    )
+
+ok <- ok  %>% sf::st_drop_geometry() %>% select(ADEP:CALL)
+nok<- nok %>% sf::st_drop_geometry() %>% select(ADEP:CALL)
+
+ds_nok_adep <- bind_rows(ok, nok) %>% append_region()
+
+ds_nok_ades <- ds_nok %>% filter(!is.na(ADEP), is.na(ADES)) %>%
+  pts_latlon_sf("LAT2","LON2") %>% 
+  geo_sfpts(eez_land) %>% 
+    mutate(ISO2 = countrycode(ISO_TER1, origin = "iso3c", destination = "iso2c")
+           ,ADES     = case_when(ALT2 <= 5000 ~ 'XLOX', ALT2 > 5000 ~ 'XHIX', TRUE ~ as.character(NA))
+           ,ADES_CTRY= case_when( ADES == "XHIX" ~ "Other"
+                                 ,ADES == "XLOX" & !is.na(ISO2) ~ ISO2
+                                 ,ADES == "XLOX" & is.na(ISO2) ~ ADEP_CTRY
+                                 ,TRUE ~ as.character(NA))
+    ) %>% 
+  sf::st_drop_geometry() %>% select(ADEP:CALL) %>% 
+  append_region()
+
+ds_nok <- bind_rows(ds_nok_adep, ds_nok_ades)   # verify number of flights
+
+#--------- combine again
+ds <- bind_rows(ds_ok, ds_nok)
+
+#-------------------
 
@@
 my_fn <- gsub(pattern = "-raw", replacement = "", x = .fn)
-my_fn <- gsub(pattern = "flightlist_", replacement = "base_osn_", x = my_fn)
+my_fn <- gsub(pattern = "flightlist_", replacement = "base_osn_2_", x = my_fn)
 my_fn <- gsub(pattern = "\\.gz", replacement = "", x = my_fn)

### 2021-07-21 x0_data-preparation.Rmd
Geschafft!!

@@
 prep_base_data <- function(.fn = file_names, ...){
+
 ds <- read_osn_covid(.fn)
@@
 ds_nok <- ds %>% filter( is.na(ADEP) | is.na(ADES))
+
+#----------------------- geo referencing of lat-lon to complement ADEP/ADES
+eez_land <- load_shapefile_for_geo_ref()
 
@@
 walk(fns, .f = prep_base_data)
+walk(fns[13:length(fns)], .f = prep_base_data)
 ```

### 2021-07-22 paper.Rmd
starting final drafting

--- 2021-07-16 paper.Rmd
@@
  #   base_format: rticles::ieee_article
-  bookdown::word_document2
+  bookdown::word_document2:
+  #   reference_docx: tmpl_Conference-template-letter.docx
+    default
 bibliography: "DASC2021.bib"
@@
 
-For several years, many concerns of the global air traffic management community has been directed to the evident problem of imbalances between capacity and demand.
+For many years, many concerns of the global air traffic management community has been directed to the evident problem of imbalances between capacity and demand.
 The pressing, increasing demand for air transport registered in the last decade not only has already produced challenging delay management practices, but also fostered projections of even worse scenarios.
@@
 As a result, air navigation providers collected less fees for their services, due to significantly fewer flights.
-In the operational perspective, pilots and air traffic controllers practiced less. The problems and obstacles cascaded into many other dimensions.
+In the operational perspective, pilots and air traffic controllers practiced less. The problems and obstacles developed into many other dimensions.
 
@@
 A variety of studies showed the interplay or consequences of the travel constraints. 
-To date, lower attention was given to the inherent change in terms of air transport services.
+To date, lower attention was given to the inherent change in terms of air transport services
 Hence, the current scenario is a proper moment to further investigate the concept of resilience.
@@
 
-* provide some descriptive text about COVID and government reactions (DONE BELOW!):
-
-For example, in March 12, 2020, the United States established the suspension of most travels from the European Schengen area (BBC, 2020).
-A few days later, European Comission President Ursula von der Leyen announced a similar measure, proposing for EU Member States to apply restrictions on non-essential travel.
-By that time, several countries, such as Austria, Germany, and Poland had already implemented border check controls, while Portugal, Italy, and Spain have imposed lockdowns (Nielsen, 2020).
-At the same time, Australia imposed a severe arrival block on all non-citizens and non-residents at its airports (Worthington & Snape, 2020).
-In South America, Brazil, Chile, Argentina and other countries also followed the trend and closed their borders, totally or partially (France24, 2020).
-As a result, an unparalleled air traffic decline was seen worldwide (ICAO, 2021).
-
-```{r, out.width='50%', fig.align='center'}
-### It's not really open data, but shows the unprecedented decline mentioned above.###
-#Use this only if you want...#
-
-knitr::include_graphics('./figures/unprecedented_decline.jpg')
-
-#Please add the references below to the bib:
-
-#BBC. (2020, March 12). Coronavirus: Trump suspends travel from Europe to US. https://www.bbc.com/news/world-us-canada-51846923
-
-#France24. (2020, March 16). Chile closes borders as Latam grounds flights over pandemic. https://www.france24.com/en/20200316-chile-closes-borders-as-latam-grounds-flights-over-pandemic
-
-#International Civil Aviation Organization.
-(2021, July 13).
-Effects of novel Coronavirus (COVID‐19) on civil aviation:Economic impact analysis.
-https://www.icao.int/sustainability/Documents/COVID-19/ICAO_Coronavirus_Econ_Impact.pdf
-
-#Nielsen, N. (2020, March 17). Commission proposes 'non-essential travel' EU entry ban. EU Observer. https://euobserver.com/coronavirus/147764
-
-#Worthington, B., & Snape, J.
-(2020, March 19).
-Australia blocks arrival of all non-citizens, non-residents in expanded coronavirus travel ban.
-ABC News.
-https://www.abc.net.au/news/2020-03-19/coronavirus-non-resident-travel-ban-australia/12071640
-
-```
-
+* provide some descriptive text about COVID and government reactions
 * any more key milestones relevant for this study, e.g. US/EU stop, Lufthansa/BA/KLM/Air France stopping to operate to China as of late January
@@
 Recently, the term is also used frequently on the political and strategic level.
-Policymakers, operational experts, and academics concur that the concept of resilience plays a major role when addressing and assessing the extent to which organisations and systems are prepared and capable to respond to and recover from disruptions.
+Policymakers, operational experts, and academes concur that the concept of resilience plays a major role when addressing and assessing the extent to which organisations and systems are prepared and capable to respond to and recover from disruptions.
 
@@
 
--   political level: The continuity and sustainability of a stable air traffic system supports and affects other dimensions of political interest, such as commercial relations, tourism industry, public health policies, among others.
-As a result, at a political level, it is of the utmost importance that the air navigation system remains functional.
-In that understanding, political acts were necessary in face of the pandemic crisis effects on aviation.
-For example, many governments injected financial support in the airline sector in order to help  them survive the worst periods.
-In addition, measures such as tax deferrals to air navigation services providers were also adopted.
-Therefore, understanding how the air navigation system can exist in a more resilient manner is certainly beneficial for many political agents.
+-   political level: ADD SOME DESCRIPTION
 
--   operational level: While efficiency, delay management, and capacity constraints were previously the main problems affecting resilience, now the concept must be broadened to an inverted scenario.
-Airlines, ANSPs, airports, and other stakeholders in the industry must remain and manage resilience not only in order to recover from disruptions caused by excessive demand, but also from insufficient demand.
-If delays are no longer the main threat, others arise, such as insufficient funding, lack of training, and traffic unpredictability.
+-   operational level: ADD SOME DESCRIPTION
 

### 2021-07-22 paper.Rmd
v0.01 ... had to move to word (old school) as rendering got a complete mess :(

@@
 library(patchwork)
+library(roll)
+library(sf)
+library(countrycode)
+library(ggnewscale)
 
@@
 The example below shows this potential approach.
+
+<<put band graph here>>
+
 ```{r}
-#-------Remember to load and run "x0_data_preparation.Rmd". The object below is produced there!--------#
+# comparator levels
+uplim  <- .80
+lowlim <- .20
+slide  <-  60
+
+tmp <- us %>% 
+  mutate( N_sm    = roll_quantile(N, width = 7, p = .50)
+         ,MED     = roll_quantile(N, width = 14, p = .50, min_obs = 7)
+         ,UP_THR  = roll_quantile(N, width = slide, p = uplim,  min_obs = 7)
+         ,LOW_THR = roll_quantile(N, width = slide, p = lowlim, min_obs = 7)
+         ,LOW_CRIT= LOW_THR >= MED
+         ,UP_CRIT = UP_THR  <= MED
+         ,STATE   = case_when( !LOW_CRIT & !UP_CRIT ~ "LEVEL"
+                              , LOW_CRIT & !UP_CRIT ~ "CRUNCH"
+                              ,!LOW_CRIT &  UP_CRIT ~ "RECOV" 
+                              ,TRUE ~ "UNKWN")
+         ,BAND    = 4000
+         ) %>%
+  filter(DATE >= lubridate::ymd("2019-10-01"))
+
+###---------Rainer, the "new_scale_colour()" kinda resets the scale, and you can use another one for the next geom_lines.
+In the example below, I left the legend.position = "bottom" (instead of "NULL"), just to see the colors------###
+Glass <- rgb(0, 0, 0, max = 255, alpha = 0, names = "transparent")
+
+ResilienceMovingBands <- tmp %>%
+  ggplot() +
+  geom_line(aes(x = DATE, y = N), colour = "grey80") +
+  geom_line(aes(x = DATE, y = N_sm, colour = THRES2, group = "RQ"), size = 2) +
+  geom_line(aes(x = DATE, y = UP_THR), color = "blue") +
+  geom_line(aes(x = DATE, y = LOW_THR), color = "red") +
+  #geom_line(aes(x = DATE, y = MED), color = "forestgreen", size = 2) +
+  #scale_color_brewer(type = "qual") +
+  new_scale_colour() +
+  geom_line(aes(x = DATE, y = BAND-100, group= "1"), colour = if_else(tmp$LOW_CRIT, "red",Glass), size = 4) +
+  geom_line(aes(x = DATE, y = BAND+16000, group= "1"), colour = if_else(tmp$UP_CRIT, "blue", Glass), size = 4) +
+  geom_line(aes(x = DATE, y = BAND-4000, group= "1", colour = STATE), size = 10) +
+  scale_colour_manual(values = c("darkred", Glass, "darkblue")) +
+  theme_minimal() +
+  theme(legend.position = "bottom")
+
 ResilienceMovingBands +

### 2021-07-22 x0_data-preparation.Rmd
v0.01 ... had to move to word (old school) as rendering got a complete mess :(

@@
 
-base_filenames <- list.files("./data/", pattern = "base_osn_*")
+base_filenames <- list.files("./data/", pattern = "base_osn_2*")
 base_fullnames <- c(paste0("./data/", base_filenames))

### 2021-07-22 paper-references.Rmd
arrrggghh ... working on references in word mode :(

---
title: "Assessing the Global COVID19 Impact on Air Transport with Open Data"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
# failed to render and I could not finally track what kills the pdf
 bookdown::pdf_book:
   base_format: rticles::ieee_article
  # bookdown::word_document2:
  # #   reference_docx: tmpl_Conference-template-letter.docx
  #   default
bibliography: "DASC2021.bib"
---

This helper document serves exclusively to get out the references from Zotero and in a paste and copy formatting.

EUROCONTROL [@statfor_growth_2018], for example,

Reported number of cases (22. Jan 2020 through 27. May 2021) [@krispin_coronavirus_2021] based on data from John Hopkins University [@JHU_2020]

global daily flight data set [@xavier_olive_2021_4893103].

RStudio [@rstudio] tools for the R language [@R]

15 minutes or more [@strohmeier_crowdsourced_2021].

PRR 2020 [@prr_2020]

who declaring covid [@who_covid_2020]

BBC 2020 [@bbc_coronavirus_2020]

ABC 2020 australia [@worthington_covid_aus_2020]

France24 corona-americas [@france24_coronavirus_2020]

ICAO [@icao_covid_impact_2021]

resilience [@resilience_engineering_2009]

glu [@gluchshenko_2012]

D1.3 [@gluchshenko_d1_res2050_2013]

koelle [@koelle_resilience_2015]

REFERENCES ============================

### 2021-07-24 paper-visualisations.Rmd
submitted version (Word & pdf), some supporting files for the final push

---
title: "Untitled"
author: "PRU"
date: "23/07/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(roll)
```

Final push analysis

- data geo-referenced to increase hits --> base_osn_2_

```{r, message=FALSE}
## load study data - as long as not outsourced to data prep
fns <- list.files(path = "./data/", pattern = "base_osn_2_", full.names = TRUE )
col_spec <- cols(
  ADEP = col_character(),
  ADEP_CTRY = col_character(),
  ADEP_REG = col_character(),
  ADES = col_character(),
  ADES_CTRY = col_character(),
  ADES_REG = col_character(),
  TYPE = col_character(),
  DATE = col_date(format = ""),
  CALL = col_character()
)
ds  <- fns %>% map_dfr(.f = ~ read_csv(.x))
```

load aircraft data base
```{r}
ac_db <- read_csv("./data-raw/aircraftDatabase.csv"
                  ,col_types = cols(.default = "c"))

#ac_types <- read_csv("./data-raw/doc8643AircraftTypes.csv")

ds <- ds %>% left_join(ac_db %>% select(icao24, registration, typecode, icaoaircrafttype), by = c("ICAO24"="icao24"))

```

## getting the meta data table

```{r}
daily_tfc <- read_csv("./data/daily_osn.csv", col_types = cols()) %>%
  #--------- correct for double counts in January 2019
  group_by(DATE) %>%
  mutate(COUNT = row_number()) %>%
  filter(COUNT == 1)

monthly_tfc <- daily_tfc %>% mutate(MOF = lubridate::floor_date(DATE, unit = "month")) %>%
  group_by(MOF) %>% summarise(N = n(), FLTS = sum(FLTS))

dwnlds <-

table_1 <- data.frame(
   FNS  = fns
  ,SIZE = file.info(fns)$size     # this is the size of our base_osn files - redo!
  ,FLTS = monthly_tfc$FLTS 
  ,YEAR = c(rep(2019, 12), rep(2020, 12), rep(2021, 6))
  )

paper_tbl <- table_1 %>%
  group_by(YEAR) %>%
  summarise(SIZE = sum(SIZE), FLTS = sum(FLTS))
```

##

```{r}
# summarise regional level
## -------- percentile thresholds
pcts   <- c(0.8, 0.6, 0.3, 0.1)

## -------- function definition
regional_timeline <- function(.ds, .reg, .pcts=pcts){
us <- .ds %>% 
  filter(ADEP_REG == .reg | ADES_REG == .reg) %>%
  group_by(YR = year(DATE), DATE) %>% 
  summarize(N = n(), .groups = "drop")

# static thresholds, x-th percentile rounded to lower integer
us_ref <- us %>% filter(YR == 2019) %>%
  pull(N) %>% quantile(probs = .pcts) %>% floor()

us_ref2 <- us %>% filter(YR == 2019, DATE >= lubridate::ymd("2019-07-01")) %>%
  pull(N) %>% quantile(probs = .pcts) %>% floor()

# use % of base level
base <- us_ref2[1]    # 80th percentile
cut_lvls <- base * c(1, 0.8, 0.6, 0.4, 0.2)
cut_lvls <- floor(cut_lvls)

max_N <- max(us$N)
lvls  <- c("Nom", "Base", "-20%", "-40%", "-60%","-80","low")
us <- us %>% 
  mutate( #THRES  = cut(N, breaks = c(max_N, us_ref, 0),  labels = lvls)
         #,
         THRES2 = cut(N, breaks = c(max_N, cut_lvls, 0)
                      #, labels = lvls
                      ,dig.lab=12    # to force full numbers
                      )
         )
return(list(us, us_ref, us_ref2, cut_lvls))
}

## do for us
res <- regional_timeline(ds, "US")
us  <- res[[1]]
us_ref2 <- res[[3]]
cut_lvls <- res[[4]]
```

```{r fig-us-timeline, fig.cap="(ref:fig-us-timeline)"}
plot_timeline <- function(.df, .pct_ref){
 p <- .df %>%
  ggplot(aes(x = DATE)) +
  geom_hline(yintercept = .pct_ref, linetype = "dotdash") +
  geom_point(aes(y = N, color = THRES2), alpha = 0.3) +
  geom_line( aes(y = N, colour = THRES2, group = 1)) +
  theme_minimal()
 return(p)
}

 gg <- plot_timeline(us, cut_lvls)

 gg 
 #+
#   coord_cartesian(ylim=c(0, 7))
```
```{r}
us_rq <- ds %>% filter(ADEP_REG == "US" | ADES_REG == "US")
us_rq %>% summarise( N = n()
                    , NOT_P = sum(!icaoaircrafttype %in% c("L1P","L2P"))
                    , NOT_PH= sum(!icaoaircrafttype %in% c("L1P","L2P","H1P", "H1T","H2T"))
)

us_rq <- us_rq %>% filter(!icaoaircrafttype %in% c("L1P","L2P","H1P", "H1T","H2T"))                                  
```

```{r}
res <- regional_timeline(us_rq, "US")
us  <- res[[1]]
cut_lvls <- res[[4]]

lbl_df <- data.frame(DATE = lubridate::ymd("2019-07-1")
                     ,LVLS= cut_lvls) %>%
  mutate(LBL = paste0(c("base", "-20%","-40%","-60%","-80%")," (", LVLS, ")")  )

gg <- plot_timeline( us %>% filter(DATE >= lubridate::ymd("2019-07-01"))
                    ,cut_lvls) 
gg +
  geom_text(lbl_df, mapping = aes(x = DATE, y = LVLS, label = LBL)
            , nudge_y = 1500, nudge_x = 30) +
  labs(x = NULL, y = NULL) +
  theme(legend.position = "none")
```

```{r}
ggsave("./figures/us-resilience-timeline.png", dpi = 320, height = 6, width = 8)
```

```{r}
check_offset  <- function(.tfc, .ref){
  df <- .tfc %>% 
    mutate(DIFF = N - .ref
           ,CUM_DIFF = if_else(DIFF >= 0, 0, DIFF)
           ,MOF = lubridate::floor_date(DATE, unit = "month"))
  return(df)
}
resilience_dt <- function(.tfc, .ref){
  df <- .tfc %>%
    check_offset(.ref) %>%
    group_by(MOF) %>%
    summarise( N_DISR = sum(DIFF < 0)
              ,FIRST  = DATE[which(DIFF < 0)[1]]
              ,CUM_DIFF = sum(CUM_DIFF))
  return(df)
}
```

```{r}
resi_us <- resilience_dt (
   us %>% filter(DATE >= lubridate::ymd("2019-07-01"))
  ,cut_lvls[2] 
)
resi
```

```{r}
bra <- ds %>% filter(ADEP_REG == "BR" | ADES_REG == "BR") %>%
  filter(!icaoaircrafttype %in% c("L1P","L2P","H1P", "H1T","H2T"))

plot_me <- function(.tfc, .reg, .nudge_y){
res <- regional_timeline(.tfc, .reg)
us  <- res[[1]]
cut_lvls <- res[[4]]

lbl_df <- data.frame(DATE = lubridate::ymd("2019-07-1")
                     ,LVLS= cut_lvls) %>%
  mutate(LBL = paste0(c("base", "-20%","-40%","-60%","-80%")," (", LVLS, ")")  )

gg_bra <- plot_timeline( us %>% filter(DATE >= lubridate::ymd("2019-07-01"))
                    ,cut_lvls) 
gg_bra <- gg_bra +
  geom_text(lbl_df, mapping = aes(x = DATE, y = LVLS, label = LBL)
            , nudge_y = .nudge_y, nudge_x = 30) +
  labs(x = NULL, y = NULL) +
  theme(legend.position = "none")
return(gg_bra)
}
```

```{r}
( g_bra <- plot_me(bra, "BR", 75) )
# ggsave("./figures/bra-resilience-timeline.png", dpi = 320, height = 6, width = 8)
```

```{r}
eur <- ds %>% filter(ADEP_REG == "EU" | ADES_REG == "EU") %>%
  filter(!icaoaircrafttype %in% c("L1P","L2P","H1P", "H1T","H2T"))

( g_eur <- plot_me(eur, "EU", 1000) )
# ggsave("./figures/eur-resilience-timeline.png", dpi = 320, height = 6, width = 8)
```

```{r, message=FALSE}
library(patchwork)
g_eur + g_bra
```

```{r}
resi_us <- resi

bra_res <- regional_timeline(bra, "BR")
resi_bra <- resilience_dt(bra_res[[1]], bra_res[[4]][2]) %>%
  filter(MOF >= lubridate::ymd("2019-07-01"))

eur_res <- regional_timeline(eur, "EU")
resi_eur <- resilience_dt(eur_res[[1]], eur_res[[4]][2]) %>%
  filter(MOF >= lubridate::ymd("2019-07-01"))
```

show this in the presentation

```{r}
aggregate_resilience <- function(.ds){
  my_agg <- .ds %>% 
    mutate(YR = lubridate::year(MOF)) %>% 
    filter(YR %in% c(2020, 2021)) %>% 
    group_by(YR) %>% 
    summarise(N_DISR = sum(N_DISR), sum(CUM_DIFF))
}

us_res <- aggregate_resilience(resi_us) %>% mutate(REG = "US")
eur_res<- aggregate_resilience(resi_eur) %>% mutate(REG= "EU")
bra_res<- aggregate_resilience(resi_bra) %>% mutate(REG= "BR")

res_all <- bind_rows(us_res, eur_res, bra_res)
res_all %>% select(REG, everything())
```

## Airport level analysis

```{r}
apt_timeline <- function(.df){
  df <- .df %>% group_by(DATE) %>%
    summarise(N_DEP = n()
              ,DEP_DOM = sum(ADEP_CTRY == ADES_CTRY)
              ,DEP_REG = sum(ADEP_REG == ADES_REG))
}

plot_timeline <- function(.df){
  df <- .df %>% 
  mutate(N_DEP_SM = roll_quantile(N_DEP, width = 7, p = .50))

  p <- df %>%  
  ggplot() +
  geom_line(aes(x = DATE, y = N_DEP), colour = "grey60") +
  geom_line(aes(x = DATE, y = N_DEP_SM)) +
  scale_x_date( limits = c(as.Date("2020-01-01"), as.Date("2021-06-01"))
               ,date_breaks = "60 days", date_labels = "%b\n%y") +
  labs(x = NULL, y = NULL) +
  theme_minimal()

  return(p)
}

dep_vs_domshare <- function(.df){
  p <- .df %>%
  ggplot() +
  geom_point(aes(y = N_DEP, x = DEP_REG / N_DEP, colour = as.factor(year(DATE)))) +
  labs(colour = NULL, x = NULL, y = NULL
       ,subtitle = "share regional departures") + 
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)
                     ,expand = c(0,0), limits = c(0,NA)) +
  theme_minimal() +
  theme(legend.position = c(0.1, 0.8))
return(p)
}

check_apt <- function(.mvts, .apt){
  apt_mvts <- .mvts %>%
    filter(ADEP == .apt) %>% 
    apt_timeline() %>%
    filter(N_DEP > 10)   # remove outlier

  p1 <- apt_mvts  %>% plot_timeline()
  p2 <- apt_mvts %>% dep_vs_domshare()

  p <- (p1 + labs(subtitle = .apt)) + p2
return(p)
}
```

```{r, message=FALSE}
apt  <- "EGLL"
mvts <- ds %>% filter(ADEP == apt) 
check_apt(mvts, apt)
ggsave(paste0("./figures/", apt, "-combi.png"), dpi = 320, height = 4, width = 8)
```

ggsave("./figures/egll-combi.png", dpi = 320, height = 4, width = 8)

```{r, message=FALSE}
apt  <- "KATL"
mvts <- ds %>% filter(ADEP == apt) 
check_apt(mvts, apt)
ggsave(paste0("./figures/", apt, "-combi.png"), dpi = 320, height = 4, width = 8)
```

```{r, message=FALSE}
apt  <- "SBGR"
mvts <- ds %>% filter(ADEP == apt) 
check_apt(mvts, apt)
ggsave(paste0("./figures/", apt, "-combi.png"), dpi = 320, height = 4, width = 8)
```

## COVID vs national level

```{r}
coronavirus::update_dataset()  # pulls latest (data)version
library(coronavirus)           # then call library

covid <- coronavirus %>% as_tibble()
```

```{r}
ctry <- "BR"

deps <- bra %>% filter(ADEP_CTRY == ctry)
deps <- deps %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarise(N_DEP = n())
```

```{r}
deps %>% ggplot() + geom_line(aes(x = DATE, y = N_DEP, group = ADES_REG, colour = ADES_REG))
 covs <- covid %>% filter(country == "Brazil") %>% filter(type == "death") %>% mutate(cum_cases = cumsum(cases)) %>% ggplot() + geom_line(aes(x = date, y = cum_cases))
```

```{r}
deps_covs <- deps %>% 
  left_join( covid %>% filter(country == "Brazil", type == "death") %>% 
              mutate(cum_cases = cumsum(cases)) %>% 
               select(DATE = date, CUM_DEATH = cum_cases)
             , by = "DATE"
             )
```

```{r}
coeff <- 500

ggplot(data = deps_covs %>% filter(DATE >= lubridate::ymd("2020-01-01"))
       , aes(x = DATE)) +
  geom_line(aes( y = N_DEP, colour = ADES_REG)) +
  geom_line(aes( y = CUM_DEATH / coeff)) +
 scale_y_continuous(
#  scale_y_log10(
    # Features of the first axis
    name = "departures",

    # Add a second axis and specify its features
    sec.axis = sec_axis( trans= ~ .*coeff, name="confirmed deaths", labels = scales::comma)
  ) +
  geom_vline(xintercept = lubridate::ymd("2020-03-15"), linetype = "dotdash") +
  labs(x = NULL, colour = NULL) +
  theme_minimal() +
  theme(legend.position = c(0.4,0.7))
```

ggsave("./figures/bra-dep-death.png", dpi = 320, width = 8, height = 5)

```{r}
deps <- eur %>% filter(ADEP_CTRY == "DE") %>% 
  group_by(DATE, ADEP_REG, ADES_REG) %>% summarise(N_DEP = n(), .groups = "drop")

deps_covs <- deps %>% 
  left_join( covid %>% filter(country == "Germany", type == "death") %>% 
              mutate(cum_cases = cumsum(cases)) %>% 
               select(DATE = date, CUM = cum_cases)
             , by = "DATE"
             )
```

```{r}
coeff <- 35

ggplot(data = deps_covs %>% filter(DATE >= lubridate::ymd("2020-01-01"))
       , aes(x = DATE)) +
  geom_line(aes( y = N_DEP, colour = ADES_REG)) +
  geom_line(aes( y = CUM / coeff)) +
 scale_y_continuous(
#  scale_y_log10(
    # Features of the first axis
    name = "departures",

    # Add a second axis and specify its features
    sec.axis = sec_axis( trans= ~ .*coeff, name="confirmed deaths", labels = scales::comma)
  ) +
  geom_vline(xintercept = lubridate::ymd("2020-03-15"), linetype = "dotdash") +
  labs(x = NULL, colour = NULL) +
  theme_minimal() +
  theme(legend.position = c(0.3,0.8))
```

ggsave("./figure/ger-dep-death.png", dpi = 320, width = 8, height = 5)
